---
title: "V01 — CUP Visium Cohort: QC & Normalisation"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 14
    fig-height: 8
execute:
  warning: false
  message: false
  cache: false
---

20 CUP Visium samples from Marta's cohort. All analyses are per-sample; no batch correction
is applied — each downstream script operates on individual samples.

**Output**: `visium_seurat_list.rds` — named list of 20 filtered, SCTransform-normalised Seurat
objects, one per sample.

```{r setup}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tibble)

options(Seurat.object.assay.version = "v5")
options(future.globals.maxSize = 4 * 1024^3)

# BASE points to the repository root.
# Each sample is expected in a sub-directory named by its sample ID,
# containing the SpaceRanger output (filtered_feature_bc_matrix.h5 + spatial/).
BASE    <- "."
OUT_RDS <- file.path(BASE, "results", "01_qc", "visium_seurat_list.rds")
```

## Sample cohort

```{r metadata}
sample_meta <- tribble(
  ~sample,   ~h5_file,
  "114B",    "filtered_feature_bc_matrix.h5",
  "271",     "filtered_feature_bc_matrix.h5",
  "274",     "filtered_feature_bc_matrix.h5",
  "297",     "filtered_feature_bc_matrix.h5",
  "351",     "filtered_feature_bc_matrix.h5",
  "353",     "filtered_feature_bc_matrix.h5",
  "355",     "filtered_feature_bc_matrix.h5",
  "368",     "filtered_feature_bc_matrix.h5",
  "369",     "filtered_feature_bc_matrix.h5",
  "372",     "filtered_feature_bc_matrix.h5",
  "373",     "filtered_feature_bc_matrix.h5",
  "374",     "filtered_feature_bc_matrix.h5",
  "375",     "filtered_feature_bc_matrix.h5",
  "376",     "filtered_feature_bc_matrix.h5",
  "381",     "filtered_feature_bc_matrix.h5",
  "383",     "filtered_feature_bc_matrix.h5",
  "385",     "filtered_feature_bc_matrix.h5",
  "386",     "filtered_feature_bc_matrix.h5",
  "387",     "filtered_feature_bc_matrix.h5",
  "394",     "filtered_feature_bc_matrix.h5"
)

knitr::kable(sample_meta, caption = "CUP Visium cohort — 20 samples")
```

## Load samples

```{r load-samples, results='asis'}
# Single cohort — no study grouping needed

load_sample <- function(s, meta) {
  row  <- meta[meta$sample == s, ]
  sdir <- file.path(BASE, s)
  h5   <- row$h5_file

  h5_path <- file.path(sdir, h5)
  counts  <- Read10X_h5(h5_path)
  if (is.list(counts)) counts <- counts[["Gene Expression"]]
  obj <- CreateSeuratObject(counts = counts, assay = "Spatial", project = s)
  img <- Read10X_Image(image.dir    = file.path(sdir, "spatial"),
                       image.name   = "tissue_lowres_image.png",
                       assay        = "Spatial", slice = s,
                       filter.matrix = TRUE)
  img <- img[Cells(obj)]
  DefaultAssay(img) <- "Spatial"
  obj[[s]] <- img

  obj$sample <- s
  obj
}

all_objects <- list()
for (s in sample_meta$sample) {
  cat(sprintf("Loading %s ... ", s))
  obj <- tryCatch(load_sample(s, sample_meta),
                  error = function(e) { cat(sprintf("ERROR: %s\n", e$message)); NULL })
  if (!is.null(obj)) {
    all_objects[[s]] <- obj
    cat(sprintf("%d spots × %d genes\n", ncol(obj), nrow(obj)))
  }
}
cat(sprintf("\nLoaded %d / %d samples\n", length(all_objects), nrow(sample_meta)))
```

## QC metrics

```{r qc-metrics}
all_objects <- lapply(all_objects, function(obj) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-", assay = "Spatial")
  obj
})
```

### Summary table

```{r qc-table}
qc_df <- lapply(names(all_objects), function(s) {
  md <- all_objects[[s]]@meta.data
  data.frame(
    sample       = s,
    n_spots      = nrow(md),
    median_UMI   = round(median(md$nCount_Spatial)),
    median_genes = round(median(md$nFeature_Spatial)),
    median_mt    = round(median(md$percent.mt), 1)
  )
}) |> bind_rows()

knitr::kable(qc_df, caption = "Per-sample QC summary — pre-filter, in-tissue spots only")
```

### Violin plots

```{r qc-violin, fig.height=10}
meta_all <- lapply(names(all_objects), function(s) {
  all_objects[[s]]@meta.data |> mutate(sample_id = s)
}) |> bind_rows()

p_umi <- ggplot(meta_all, aes(sample_id, nCount_Spatial, fill = sample_id)) +
  geom_violin(scale = "width", linewidth = 0.3, alpha = 0.85) +
  geom_boxplot(width = 0.08, outlier.size = 0.3, fill = "white") +
  scale_y_log10() +
  labs(title = "UMI per spot", x = NULL, y = "nCount (log10)") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p_gene <- ggplot(meta_all, aes(sample_id, nFeature_Spatial, fill = sample_id)) +
  geom_violin(scale = "width", linewidth = 0.3, alpha = 0.85) +
  geom_boxplot(width = 0.08, outlier.size = 0.3, fill = "white") +
  labs(title = "Genes per spot", x = NULL, y = "nFeature") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p_mt <- ggplot(meta_all, aes(sample_id, percent.mt, fill = sample_id)) +
  geom_violin(scale = "width", linewidth = 0.3, alpha = 0.85) +
  geom_boxplot(width = 0.08, outlier.size = 0.3, fill = "white") +
  labs(title = "Mitochondrial %", x = NULL, y = "% MT") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

p_umi / p_gene / p_mt
```

### Spatial QC maps

```{r qc-spatial, fig.height=5, results='asis'}
for (s in names(all_objects)) {
  cat(sprintf("\n#### %s\n\n", s))
  print(
    SpatialFeaturePlot(all_objects[[s]],
                       features       = c("nCount_Spatial", "nFeature_Spatial", "percent.mt"),
                       ncol           = 3,
                       pt.size.factor = 1.4,
                       image.alpha    = 0.35) &
      theme(legend.position = "right")
  )
  cat("\n")
}
```

## Filter

Thresholds applied uniformly across all samples. Adjust if any sample has an unusual UMI
distribution (inspect violin plots above before rendering the report).

```{r filter}
QC_MIN_UMI   <- 500
QC_MAX_UMI   <- 50000
QC_MIN_GENES <- 300
QC_MAX_MT    <- 20

filtered <- lapply(all_objects, function(obj) {
  subset(obj,
    nCount_Spatial   >= QC_MIN_UMI   &
    nCount_Spatial   <= QC_MAX_UMI   &
    nFeature_Spatial >= QC_MIN_GENES &
    percent.mt       <= QC_MAX_MT)
})

filt_stats <- data.frame(
  sample   = names(all_objects),
  before   = sapply(all_objects, ncol),
  after    = sapply(filtered, ncol)
) |> mutate(removed = before - after, pct_kept = round(100 * after / before, 1))

knitr::kable(filt_stats, caption = "Spots retained after QC filtering")
```

## Normalisation — SCTransform per sample

Each sample is normalised independently with SCTransform. The resulting `SCT` assay is used
for dimensionality reduction; raw counts in `Spatial` assay are preserved for CancerFinder3
export in V03.

```{r normalise}
filtered <- parallel::mclapply(filtered, function(obj) {
  SCTransform(obj, assay = "Spatial", verbose = FALSE)
}, mc.cores = 14)
```

## Overview UMAP (QC only — no batch correction)

This merged embedding is for quality inspection only. All downstream analyses are per-sample.

```{r overview-umap, fig.height=6}
merged <- merge(filtered[[1]], filtered[-1], add.cell.ids = names(filtered))
DefaultAssay(merged) <- "SCT"

hvg_common <- Reduce(intersect, lapply(filtered, VariableFeatures))
cat(sprintf("Common HVGs across all samples: %d\n", length(hvg_common)))
VariableFeatures(merged) <- hvg_common

merged <- RunPCA(merged, npcs = 30, verbose = FALSE)
merged <- RunUMAP(merged, dims = 1:20, verbose = FALSE)

p_sample <- DimPlot(merged, group.by = "sample", pt.size = 0.3) + ggtitle("Sample")

p_sample
```

## Save checkpoint

```{r save}
saveRDS(filtered, OUT_RDS)
cat(sprintf("Saved %d samples → %s\n", length(filtered), OUT_RDS))
lapply(names(filtered), function(s) {
  cat(sprintf("  %s: %d spots\n", s, ncol(filtered[[s]]))  )
})
```

## Session info

```{r session}
sessionInfo()
```
