---
title: "V02 — BayesSpace Spatial Enhancement"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 14
    fig-height: 8
execute:
  warning: false
  message: false
  cache: false
---

Runs BayesSpace per-sample to (1) identify spatially coherent tissue domains (`spatialCluster`)
and (2) generate enhanced sub-spot pseudo-cells (`spatialEnhance`).

**Resolution**: `nsubspot.per.edge = 4` subdivides each Visium spot (~55 µm, ~100 µm pitch)
into ~13 sub-spots, giving an effective sub-spot spacing of ~25 µm — sufficient to resolve the
<30 µm Tumour niche zone used in V06.

**Parallelisation**: samples are processed in parallel with `mclapply`. BayesSpace's MCMC
(Gibbs sampling) runs on CPU only — no GPU path exists in the R package. The GPU can be used
in V03 for CancerFinder3's PyTorch inference instead. Remaining cores are allocated to
OpenBLAS (BLAS threads) to accelerate the PCA step within each worker.

**Runtime**: with `N_PARALLEL = 5` and `nrep = 10 000`, expect ~60–90 min total (all 20
samples). Per-sample RDS caches are written to `bayesspace_cache/`; re-running skips
completed samples.

**Input**: `visium_seurat_list.rds`
**Output**: `visium_enhanced_sce_list.rds` + `bayesspace_cache/<sample>_enhanced.rds`

```{r setup}
library(Seurat)
library(BayesSpace)
library(SingleCellExperiment)
library(Matrix)
library(parallel)
library(ggplot2)
library(dplyr)
library(patchwork)
library(scales)

BASE      <- "."
CACHE_DIR <- file.path(BASE, "bayesspace_cache")
OUT_RDS   <- file.path(BASE, "visium_enhanced_sce_list.rds")
dir.create(CACHE_DIR, showWarnings = FALSE)

# ── Parallelisation config ──────────────────────────────────────────────────
# Run up to 5 samples in parallel (reduced for USB drive I/O; 20 samples = two passes)
# 24 usable cores split across workers:
#   BLAS_PER_WORKER  = 2  (OpenBLAS threads for PCA in spatialPreprocess)
#   ENHANCE_CORES    = 2  (cores argument to spatialEnhance, parallel sub-chains)
#   → 10 × 2 = 20 cores busy during enhancement; headroom for OS
N_PARALLEL           <- 5L
TOTAL_CORES          <- as.integer(parallel::detectCores() - 2L)
BLAS_PER_WORKER      <- 2L
ENHANCE_CORES        <- 2L

# ── BayesSpace config ───────────────────────────────────────────────────────
NSUBSPOT_PER_EDGE <- 4L   # ~13 sub-spots/spot, ~25 µm effective spacing
NREP_CLUSTER      <- 10000L
NREP_ENHANCE      <- 10000L
BURN_IN           <- 1000L

seurat_list <- readRDS(file.path(BASE, "results", "01_qc", "visium_seurat_list.rds"))

SAMPLES_SKIP <- c()   # no samples to skip
if (length(SAMPLES_SKIP) > 0)
  seurat_list <- seurat_list[setdiff(names(seurat_list), SAMPLES_SKIP)]
```

## Helper functions

```{r helpers}
make_bayesspace_sce <- function(seurat_obj, sample_name) {
  counts_mat <- GetAssayData(seurat_obj, assay = "Spatial", layer = "counts")
  tp_path <- file.path(BASE, sample_name, "spatial", "tissue_positions.csv")
  tp <- read.csv(tp_path, row.names = 1)
  cells <- intersect(colnames(seurat_obj), rownames(tp))
  tp <- tp[cells, ]

  sce <- SingleCellExperiment(
    assays  = list(counts = counts_mat[, cells, drop = FALSE]),
    colData = DataFrame(
      row.names          = cells,
      spot_id            = cells,
      array_row          = tp$array_row,
      array_col          = tp$array_col,
      pxl_row_in_fullres = tp$pxl_row_in_fullres,
      pxl_col_in_fullres = tp$pxl_col_in_fullres
    )
  )
  sce
}
```

## Run BayesSpace — parallel across samples

Each sample is processed in an independent forked worker. Libraries and the full R environment
are inherited via `fork()` — no re-loading needed. BLAS threads are set per worker to use the
remaining cores for the PCA step in `spatialPreprocess`.

```{r run-bayesspace}
process_sample <- function(s) {
  cache_file <- file.path(CACHE_DIR, paste0(s, "_enhanced.rds"))
  fail_file  <- file.path(CACHE_DIR, paste0(s, "_failed.flag"))

  if (file.exists(cache_file)) {
    message(sprintf("[%s] cache hit — skipping", s))
    return(invisible(NULL))
  }

  # Top-level error guard: any failure writes a .flag and returns NULL (no worker crash)
  tryCatch({
    if (requireNamespace("RhpcBLASctl", quietly = TRUE))
      RhpcBLASctl::blas_set_num_threads(BLAS_PER_WORKER)

    message(sprintf("[%s] starting (pid %d)", s, Sys.getpid()))
    t0 <- proc.time()

    sce <- make_bayesspace_sce(seurat_list[[s]], s)
    message(sprintf("[%s] %d spots × %d genes", s, ncol(sce), nrow(sce)))

    set.seed(42)
    sce <- spatialPreprocess(sce, platform = "Visium",
                             n.PCs = 15, n.HVGs = 2000,
                             log.normalize = TRUE, skip.PCA = FALSE)

    set.seed(42)
    sce       <- qTune(sce, qs = 4:8, platform = "Visium", nrep = 1000)
    q_logliks <- attr(sce, "q.logliks")
    q_opt     <- q_logliks$q[which.max(q_logliks$loglik)]
    message(sprintf("[%s] optimal q = %d  (log-liks: %s)",
                    s, q_opt,
                    paste(sprintf("q%d=%.0f", q_logliks$q, q_logliks$loglik), collapse = ", ")))

    set.seed(42)
    sce <- spatialCluster(sce, q = q_opt, platform = "Visium",
                          nrep = NREP_CLUSTER, burn.in = BURN_IN)

    # Try nsubspots.per.edge=4, fall back to 3
    set.seed(42)
    sce_enh <- tryCatch(
      spatialEnhance(sce, q = q_opt, platform = "Visium",
                     nrep               = NREP_ENHANCE,
                     burn.in            = BURN_IN,
                     nsubspots.per.edge = NSUBSPOT_PER_EDGE,
                     cores              = ENHANCE_CORES,
                     save.chain         = FALSE),
      error = function(e) {
        message(sprintf("[%s] nsubspots.per.edge=%d failed (%s) — retrying with 3",
                        s, NSUBSPOT_PER_EDGE, conditionMessage(e)))
        set.seed(42)
        spatialEnhance(sce, q = q_opt, platform = "Visium",
                       nrep               = NREP_ENHANCE,
                       burn.in            = BURN_IN,
                       nsubspots.per.edge = 3L,
                       cores              = ENHANCE_CORES,
                       save.chain         = FALSE)
      }
    )

    parent_idx <- colData(sce_enh)$spot.idx
    colData(sce_enh)$pxl_row_in_fullres  <- colData(sce)$pxl_row_in_fullres[parent_idx]
    colData(sce_enh)$pxl_col_in_fullres  <- colData(sce)$pxl_col_in_fullres[parent_idx]
    colData(sce_enh)$original_spot_id    <- colnames(sce)[parent_idx]
    colData(sce_enh)$parent_array_row    <- colData(sce)$array_row[parent_idx]
    colData(sce_enh)$parent_array_col    <- colData(sce)$array_col[parent_idx]
    colData(sce_enh)$sample              <- s

    elapsed <- round((proc.time() - t0)[["elapsed"]] / 60, 1)
    message(sprintf("[%s] done — %d sub-spots in %.1f min; saving ...", s, ncol(sce_enh), elapsed))
    saveRDS(sce_enh, cache_file)

  }, error = function(e) {
    msg <- conditionMessage(e)
    message(sprintf("[%s] FAILED — %s", s, msg))
    writeLines(msg, fail_file)   # persist failure reason for inspection
  })

  invisible(NULL)
}

# Run N_PARALLEL samples simultaneously
results <- mclapply(names(seurat_list), process_sample,
                    mc.cores = N_PARALLEL, mc.preschedule = FALSE)
names(results) <- names(seurat_list)

# Report failures (warn, do not stop — failed samples are skipped downstream)
fail_flags <- list.files(CACHE_DIR, pattern = "_failed\\.flag$", full.names = TRUE)
if (length(fail_flags) > 0) {
  failed_names <- sub("_failed\\.flag$", "", basename(fail_flags))
  for (fn in fail_flags) {
    reason <- tryCatch(readLines(fn, warn = FALSE), error = function(e) "unknown error")
    warning(sprintf("BayesSpace SKIPPED %s: %s", sub("_failed\\.flag$", "", basename(fn)),
                    paste(reason, collapse = " ")))
  }
  cat(sprintf("\n> **%d sample(s) skipped** (BayesSpace failed): %s\n\n",
              length(failed_names), paste(failed_names, collapse = ", ")))
} else {
  cat("All samples complete.\n")
}
```

## Load enhanced SCEs from cache

`spatialEnhance` returns sub-spot positions and cluster labels only (no assays). We map
parent-spot raw counts → log-normalise → assign to each sub-spot using the stored
`original_spot_id` field, giving downstream scripts a `logcounts` assay to work with.

```{r load-all}
enhanced_list <- lapply(names(seurat_list), function(s) {
  f <- file.path(CACHE_DIR, paste0(s, "_enhanced.rds"))
  if (!file.exists(f)) {
    message(sprintf("[%s] no cache — skipped (BayesSpace failed or not yet run)", s))
    return(NULL)
  }
  sce_enh <- readRDS(f)

  # spatialEnhance produces no assay data; populate logcounts from parent-spot raw counts.
  raw        <- GetAssayData(seurat_list[[s]], assay = "Spatial", layer = "counts")
  parent_ids <- colData(sce_enh)$original_spot_id   # parent barcode per sub-spot

  # Guard: replace any NA parent_ids (edge case for last spot's sub-spots)
  na_mask <- is.na(parent_ids)
  if (any(na_mask)) {
    parent_ids[na_mask] <- colnames(seurat_list[[s]])[1]
    message(sprintf("[%s] %d sub-spots had NA parent_id — assigned to first spot", s, sum(na_mask)))
  }

  # Log-normalise raw counts for the selected parents (columns = one entry per sub-spot)
  parent_raw  <- raw[, parent_ids, drop = FALSE]
  lib_sizes   <- Matrix::colSums(parent_raw)
  lib_sizes[lib_sizes == 0] <- 1L
  parent_lc   <- log1p(Matrix::t(Matrix::t(parent_raw) / lib_sizes) * 10000)
  colnames(parent_lc) <- colnames(sce_enh)

  assay(sce_enh, "logcounts") <- parent_lc
  sce_enh
})
names(enhanced_list) <- names(seurat_list)
enhanced_list <- Filter(Negate(is.null), enhanced_list)
cat(sprintf("Loaded %d / %d enhanced SCEs\n", length(enhanced_list), length(seurat_list)))
```

## Enhanced sub-spot QC

Sub-spots inherit parent expression (BayesSpace `spatialEnhance` does not deconvolve
expression — positions and cluster labels are enhanced). QC is therefore reported at both
levels: **parent spots** (true observation units for expression) and **sub-spots** (spatial
resolution units).

```{r qc-build}
qc_df <- lapply(names(enhanced_list), function(s) {
  sce_enh <- enhanced_list[[s]]
  lc      <- assay(sce_enh, "logcounts")   # genes × sub-spots

  # ── Sub-spot level ──────────────────────────────────────────────────────────
  subspot_counts  <- Matrix::colSums(lc)                          # total logcount sum
  subspot_genes   <- Matrix::colSums(lc > 0)                      # detected genes
  subspot_zeros   <- 1 - subspot_genes / nrow(lc)                 # sparsity fraction

  # ── Parent-spot level (from seurat raw counts) ─────────────────────────────
  meta        <- seurat_list[[s]]@meta.data
  parent_ids  <- colData(sce_enh)$original_spot_id
  n_subspots  <- as.integer(table(parent_ids)[unique(parent_ids)])

  # Align parent metadata to the unique parent order
  parent_meta <- meta[unique(parent_ids), , drop = FALSE]

  data.frame(
    sample         = s,
    # sub-spot rows
    subspot_id     = colnames(sce_enh),
    subspot_counts = subspot_counts,
    subspot_genes  = subspot_genes,
    subspot_zeros  = subspot_zeros,
    parent_id      = parent_ids,
    # parent info broadcast to sub-spots (for joined plots)
    parent_umi     = meta[parent_ids, "nCount_Spatial"],
    parent_genes   = meta[parent_ids, "nFeature_Spatial"],
    n_subspots_for_parent = as.integer(table(parent_ids)[parent_ids])
  )
}) |> bind_rows()
```

### Sub-spots per parent spot

```{r qc-subspots-per-parent, fig.height=4}
# One row per unique parent spot
parent_df <- qc_df[!duplicated(paste(qc_df$sample, qc_df$parent_id)), ]

ggplot(parent_df, aes(x = sample, y = n_subspots_for_parent, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.5, fill = "white", show.legend = FALSE) +
  labs(title = "Sub-spots generated per parent spot",
       x = NULL, y = "Sub-spots per parent spot") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### UMI counts — parent spots vs sub-spots

```{r qc-counts, fig.height=4}
p_parent <- ggplot(parent_df, aes(x = sample, y = parent_umi, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  scale_y_log10() +
  labs(title = "Raw UMI per parent spot", x = NULL, y = "nCount (log10)") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_sub <- ggplot(qc_df, aes(x = sample, y = subspot_counts, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  labs(title = "Total logcount sum per sub-spot", x = NULL, y = "Sum of logcounts") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_parent | p_sub
```

### Genes detected — parent spots vs sub-spots

```{r qc-genes, fig.height=4}
p_pgenes <- ggplot(parent_df, aes(x = sample, y = parent_genes, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  labs(title = "Genes detected per parent spot", x = NULL, y = "nFeature") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_sgenes <- ggplot(qc_df, aes(x = sample, y = subspot_genes, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  labs(title = "Genes detected per sub-spot", x = NULL, y = "nFeature") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_pgenes | p_sgenes
```

### Sparsity (zero fraction) per sub-spot

```{r qc-sparsity, fig.height=4}
ggplot(qc_df, aes(x = sample, y = subspot_zeros, fill = sample)) +
  geom_violin(scale = "width", linewidth = 0.3, show.legend = FALSE) +
  geom_boxplot(width = 0.06, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Fraction of zeros per sub-spot (logcounts == 0)",
       x = NULL, y = "Zero fraction") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### QC summary table

```{r qc-table}
qc_summary <- lapply(names(enhanced_list), function(s) {
  sq  <- qc_df[qc_df$sample == s, ]
  pq  <- sq[!duplicated(sq$parent_id), ]
  data.frame(
    sample              = s,
    n_parent_spots      = nrow(pq),
    n_subspots          = nrow(sq),
    median_subspots_per_parent = round(median(pq$n_subspots_for_parent), 1),
    median_parent_umi   = round(median(pq$parent_umi)),
    median_parent_genes = round(median(pq$parent_genes)),
    median_subspot_genes = round(median(sq$subspot_genes)),
    median_zero_frac    = round(median(sq$subspot_zeros), 3)
  )
}) |> bind_rows()

knitr::kable(qc_summary,
             col.names = c("Sample", "Parent spots", "Sub-spots",
                           "Median sub-spots/parent", "Median UMI/parent",
                           "Median genes/parent", "Median genes/sub-spot",
                           "Median zero frac"),
             caption = "Enhancement QC — parent-spot and sub-spot level metrics")
```

## Spatial cluster plots

```{r cluster-plots, fig.height=5, results='asis'}
for (s in names(enhanced_list)) {
  cat(sprintf("\n#### %s\n\n", s))
  sce_enh <- enhanced_list[[s]]
  coords  <- as.data.frame(colData(sce_enh))
  coords$cluster <- factor(sce_enh$spatial.cluster)

  p <- ggplot(coords, aes(array_col, -array_row, colour = cluster)) +
    geom_point(size = 0.2, alpha = 0.7) +
    coord_fixed() +
    scale_colour_brewer(palette = "Set2") +
    labs(title = s,
         subtitle = sprintf("%d sub-spots from %d spots | nsubspot.per.edge = %d",
                            nrow(coords),
                            length(unique(coords$original_spot_id)),
                            NSUBSPOT_PER_EDGE),
         x = "array_col", y = "array_row", colour = "domain") +
    theme_classic(base_size = 11) +
    guides(colour = guide_legend(override.aes = list(size = 2)))

  print(p)
  cat("\n")
}
```

## Summary

```{r summary-table}
summary_df <- lapply(names(enhanced_list), function(s) {
  sce_enh <- enhanced_list[[s]]
  data.frame(
    sample           = s,
    orig_spots       = length(unique(colData(sce_enh)$original_spot_id)),
    sub_spots        = ncol(sce_enh),
    subspots_per_spot = round(ncol(sce_enh) /
                              length(unique(colData(sce_enh)$original_spot_id)), 1),
    n_domains        = length(unique(sce_enh$spatial.cluster))
  )
}) |> bind_rows()

knitr::kable(summary_df,
             caption = sprintf("BayesSpace enhancement — nsubspot.per.edge = %d (~25 µm spacing)",
                               NSUBSPOT_PER_EDGE))
```

## Save checkpoint

```{r save}
saveRDS(enhanced_list, OUT_RDS)
cat(sprintf("Saved %d enhanced SCEs → %s\n", length(enhanced_list), OUT_RDS))
```

## Session info

```{r session}
sessionInfo()
```
