---
title: "V03 — CancerFinder3 Tumour/Stroma Annotation"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 14
    fig-height: 6
execute:
  warning: false
  message: false
  cache: false
---

CancerFinder3 is run directly on the **BayesSpace enhanced sub-spots** for consistency with all
downstream analyses. The enhanced `logcounts` are converted back to pseudo-counts via `expm1()`
before export, so that CF3 can apply its standard log-normalisation internally and receive
data in the expected scale.

**Threshold**: cancer_prob ≥ 0.85 → Tumour (same as Xenium pipeline).

**Input**: `visium_enhanced_sce_list.rds`
**Output**: `visium_annotated_sce_list.rds`, `cf3_results/<sample>_predictions.csv`

```{r setup}
library(SingleCellExperiment)
library(Matrix)
library(ggplot2)
library(dplyr)

BASE        <- "."
CF3_INPUT   <- file.path(BASE, "cf3_input")
CF3_RESULTS <- file.path(BASE, "cf3_results")
OUT_RDS     <- file.path(BASE, "visium_annotated_sce_list.rds")

# ── CancerFinder3 paths ────────────────────────────────────────────────────
# Download CancerFinder3 from https://github.com/JiangBioLab/CancerFinder
# and set the paths below to the local installation.
# Path to CancerFinder3 pre-trained model (download from https://github.com/JiangBioLab/CancerFinder)
CF3_MODEL  <- "path/to/CancerFinder3/models/st_pretrain_article.pkl"
CF3_INFER  <- "path/to/CancerFinder3/infer.py"
CF3_THRESH <- 0.85

dir.create(CF3_INPUT,   showWarnings = FALSE, recursive = TRUE)
dir.create(CF3_RESULTS, showWarnings = FALSE, recursive = TRUE)

enhanced_list <- readRDS(file.path(BASE, "visium_enhanced_sce_list.rds"))
```

## Python helper — MTX → h5ad + infer

Written once; called for each sample.

```{r write-py-helper}
py_script <- file.path(CF3_INPUT, "mtx_to_h5ad_and_infer.py")
writeLines(c(
  "import sys, os, subprocess",
  "import scipy.io",
  "import numpy as np, pandas as pd, anndata as ad",
  "",
  "sample_dir, pred_csv, model_ckp, infer_py = sys.argv[1:5]",
  "",
  "# MTX is genes × cells (standard writeMM output); transpose to cells × genes for AnnData",
  "mat      = scipy.io.mmread(os.path.join(sample_dir, 'matrix.mtx')).T.tocsr()",
  "barcodes = pd.read_csv(os.path.join(sample_dir, 'barcodes.tsv'), header=None)[0].tolist()",
  "genes    = pd.read_csv(os.path.join(sample_dir, 'genes.tsv'),    header=None)[0].tolist()",
  "",
  "adata           = ad.AnnData(X=mat)   # cells × genes",
  "adata.obs_names = barcodes",
  "adata.var_names = genes",
  "# CF3 InferLoaders expects genes × cells (obs=genes, var=cells)",
  "adata_T         = adata.T",
  "h5ad_path       = os.path.join(sample_dir, 'input_transposed.h5ad')",
  "adata_T.write_h5ad(h5ad_path)",
  "",
  "subprocess.run([sys.executable, infer_py,",
  "    '--ckp', model_ckp, '--matrix', h5ad_path,",
  "    '--out', pred_csv, '--threshold', '0.85'], check=True)"
), py_script)
```

## Export enhanced sub-spots and run CF3

```{r run-cf3, results='asis'}
for (s in names(enhanced_list)) {
  pred_csv <- file.path(CF3_RESULTS, paste0(s, "_predictions.csv"))

  if (file.exists(pred_csv)) {
    cat(sprintf("\n**%s**: CF3 predictions already exist — skipping\n\n", s))
    next
  }

  cat(sprintf("\n---\n\n### %s\n\n", s))
  sce_enh <- enhanced_list[[s]]

  # Convert logcounts → pseudo-counts (genes × sub-spots, standard SCE orientation)
  logcounts_mat <- assay(sce_enh, "logcounts")
  pseudocounts  <- Matrix::Matrix(expm1(as.matrix(logcounts_mat)), sparse = TRUE)

  sdir <- file.path(CF3_INPUT, s)
  dir.create(sdir, showWarnings = FALSE)
  Matrix::writeMM(pseudocounts, file.path(sdir, "matrix.mtx"))  # genes × sub-spots
  writeLines(rownames(sce_enh),  file.path(sdir, "genes.tsv"))
  writeLines(colnames(sce_enh),  file.path(sdir, "barcodes.tsv"))

  cat(sprintf("  Exported: %d genes × %d sub-spots\n\n", nrow(sce_enh), ncol(sce_enh)))

  cmd <- sprintf(
    "conda run -n scf python %s %s %s %s %s",
    shQuote(py_script), shQuote(sdir), shQuote(pred_csv),
    shQuote(CF3_MODEL), shQuote(CF3_INFER)
  )
  exit_code <- system(cmd)
  if (exit_code != 0) {
    warning(sprintf("CF3 failed for %s (exit code %d) — skipping", s, exit_code))
    cat(sprintf("  **SKIPPED** (CF3 failed)\n\n"))
    next
  }
  cat(sprintf("  CF3 complete → %s\n\n", pred_csv))
}
```

## Import predictions and add compartment labels

```{r import-predictions, results='asis'}
cf3_failed <- character(0)
for (s in names(enhanced_list)) {
  pred_csv <- file.path(CF3_RESULTS, paste0(s, "_predictions.csv"))
  if (!file.exists(pred_csv)) {
    cf3_failed <- c(cf3_failed, s)
    cat(sprintf("**%s**: no predictions (CF3 failed) — excluded from output\n\n", s))
    next
  }
  pred     <- read.csv(pred_csv, stringsAsFactors = FALSE)

  # infer.py writes: unnamed index | sample (cell/subspot ID) | predict (cancer prob)
  rownames(pred) <- pred$sample

  sce_enh <- enhanced_list[[s]]
  colData(sce_enh)$cancer_prob <- pred[colnames(sce_enh), "predict"]
  colData(sce_enh)$compartment <- ifelse(
    colData(sce_enh)$cancer_prob >= CF3_THRESH, "Tumour", "Stroma"
  )
  enhanced_list[[s]] <- sce_enh

  n_tumour <- sum(colData(sce_enh)$compartment == "Tumour", na.rm = TRUE)
  n_total  <- ncol(sce_enh)
  cat(sprintf("**%s**: %d / %d sub-spots = %.1f%% Tumour\n\n",
              s, n_tumour, n_total, 100 * n_tumour / n_total))
}
```

## Spatial compartment maps

```{r compartment-maps, fig.height=4.5, results='asis'}
comp_colours <- c("Tumour" = "#C0392B", "Stroma" = "#2980B9")

for (s in names(enhanced_list)) {
  cat(sprintf("\n#### %s\n\n", s))
  sce_enh <- enhanced_list[[s]]

  df <- data.frame(
    col         = colData(sce_enh)$array_col,
    row         = colData(sce_enh)$array_row,
    compartment = colData(sce_enh)$compartment
  )

  p <- ggplot(df, aes(col, -row, colour = compartment)) +
    geom_point(size = 0.25, alpha = 0.7) +
    coord_fixed() +
    scale_colour_manual(values = comp_colours) +
    labs(title = s, colour = NULL) +
    theme_classic(base_size = 11) +
    guides(colour = guide_legend(override.aes = list(size = 2.5)))

  print(p)
  cat("\n")
}
```

## Summary

```{r summary-table}
summary_df <- lapply(names(enhanced_list), function(s) {
  comp <- colData(enhanced_list[[s]])$compartment
  data.frame(
    sample     = s,
    n_subspots = length(comp),
    pct_tumour = round(100 * mean(comp == "Tumour", na.rm = TRUE), 1),
    pct_stroma = round(100 * mean(comp == "Stroma", na.rm = TRUE), 1)
  )
}) |> bind_rows()

knitr::kable(summary_df, caption = "Compartment composition per sample (enhanced sub-spots)")
```

## Save checkpoint

```{r save}
if (length(cf3_failed) > 0) {
  cat(sprintf("Excluding %d CF3-failed sample(s): %s\n",
              length(cf3_failed), paste(cf3_failed, collapse = ", ")))
  enhanced_list <- enhanced_list[setdiff(names(enhanced_list), cf3_failed)]
}
saveRDS(enhanced_list, OUT_RDS)
cat(sprintf("Saved %d annotated enhanced SCEs → %s\n", length(enhanced_list), OUT_RDS))
```

## Session info

```{r session}
sessionInfo()
```
