---
title: "V05 — PROGENy Pathway Analysis & ONC/IMI Classification"
author: "Abu Saadat"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: "right"
    toc-title: "Contents"
    toc-expand: true
    code-fold: true
    code-tools: false
    code-summary: "Hide code"
    lightbox: true
    self-contained: true
    fig-width: 12
    fig-height: 7
    fig-dpi: 150
    cap-location: bottom
    fontsize: 1.05em
    linestretch: 1.6
    theme:
      light: "cosmo"
      dark : "cyborg"
execute:
  warning: false
  message: false
  echo: false
---

Pseudobulk PROGENy (7 pathways, 500 top genes) on enhanced sub-spots, following the same
classification logic as the CUP Xenium pipeline (Script 07, v3_mixed):

- **ONC score** = mean(PI3K, EGFR, MAPK, VEGF) from **Tumour** pseudobulk
- **IMI score** = mean(JAK-STAT, NFkB, TNFα) from **Stroma** pseudobulk
- **δ = ONC − IMI**: δ > 0 → ONC; δ ≤ 0 → IMI

Scoring each pathway family in its biologically relevant compartment mirrors the Xenium
approach and allows direct comparison between CUP and known-primary metastases.

**Input**: `visium_celltypes_sce_list.rds`
**Output**: `visium_progeny_classification.csv`, pathway heatmap

```{r setup}
library(SingleCellExperiment)
library(progeny)
library(decoupleR)
library(ggplot2)
library(pheatmap)
library(dplyr)
library(tibble)
library(tidyr)

BASE <- "."
OUT_CSV <- file.path(BASE, "visium_progeny_classification.csv")

enhanced_list <- readRDS(file.path(BASE, "visium_celltypes_sce_list.rds"))

PATHWAYS_ONC <- c("PI3K", "EGFR", "MAPK", "VEGF")
PATHWAYS_IMI <- c("JAK-STAT", "NFkB", "TNFa")
ALL_PATHWAYS <- c(PATHWAYS_ONC, PATHWAYS_IMI)
```

## Build pseudobulk expression per sample × compartment

Sum logcounts across all enhanced sub-spots within each compartment (Tumour / Stroma).
Using logcounts preserves relative expression differences needed by PROGENy.

```{r pseudobulk}
pseudobulk_list <- lapply(names(enhanced_list), function(s) {
  sce_enh <- enhanced_list[[s]]
  lc      <- assay(sce_enh, "logcounts")
  comp    <- colData(sce_enh)$compartment

  tumour_idx <- which(comp == "Tumour")
  stroma_idx <- which(comp == "Stroma")

  pb_tumour <- rowSums(lc[, tumour_idx, drop = FALSE])
  pb_stroma <- rowSums(lc[, stroma_idx, drop = FALSE])
  pb_total   <- rowSums(lc)

  data.frame(
    gene           = rownames(sce_enh),
    tumour_counts  = pb_tumour,
    stroma_counts  = pb_stroma,
    total_counts   = pb_total,
    stringsAsFactors = FALSE
  )
})
names(pseudobulk_list) <- names(enhanced_list)

cat(sprintf("Pseudobulk built for %d samples\n", length(pseudobulk_list)))
```

## Run PROGENy

PROGENy is run separately on Tumour and Stroma pseudobulks. The model matrix is retrieved
from `progeny::getModel()` for the 7 selected pathways (500 top genes each).

```{r run-progeny}
# Build expression matrices: genes × samples
# Use intersection of genes across all samples (panels differ between datasets)
all_genes <- Reduce(intersect, lapply(pseudobulk_list, function(x) x$gene))
cat(sprintf("Common genes across all samples: %d\n", length(all_genes)))

tumour_mat <- sapply(pseudobulk_list, function(x) {
  vals <- x$tumour_counts; names(vals) <- x$gene; vals[all_genes]
})
stroma_mat <- sapply(pseudobulk_list, function(x) {
  vals <- x$stroma_counts; names(vals) <- x$gene; vals[all_genes]
})
rownames(tumour_mat) <- all_genes
rownames(stroma_mat) <- all_genes

# PROGENy scores (samples × pathways)
progeny_tumour <- progeny(
  tumour_mat,
  scale        = TRUE,
  organism     = "Human",
  top          = 500,
  perm         = 1,
  z_scores     = TRUE
) |> as.data.frame()

progeny_stroma <- progeny(
  stroma_mat,
  scale        = TRUE,
  organism     = "Human",
  top          = 500,
  perm         = 1,
  z_scores     = TRUE
) |> as.data.frame()

# Keep only the 7 selected pathways
pathways_avail <- intersect(ALL_PATHWAYS, colnames(progeny_tumour))
if (length(pathways_avail) < length(ALL_PATHWAYS)) {
  missing <- setdiff(ALL_PATHWAYS, pathways_avail)
  warning(sprintf("Pathways not found in PROGENy model: %s", paste(missing, collapse=", ")))
}

progeny_tumour <- progeny_tumour[, intersect(ALL_PATHWAYS, colnames(progeny_tumour)), drop=FALSE]
progeny_stroma <- progeny_stroma[, intersect(ALL_PATHWAYS, colnames(progeny_stroma)), drop=FALSE]

cat("PROGENy complete.\n")
cat(sprintf("  Tumour pathways: %s\n", paste(colnames(progeny_tumour), collapse=", ")))
cat(sprintf("  Stroma pathways: %s\n", paste(colnames(progeny_stroma), collapse=", ")))
```

## ONC / IMI scoring and classification

```{r classify}

samples <- names(enhanced_list)

# ONC score: mean of oncogenic pathways from Tumour pseudobulk
onc_pathways_found <- intersect(PATHWAYS_ONC, colnames(progeny_tumour))
imi_pathways_found <- intersect(PATHWAYS_IMI, colnames(progeny_stroma))

onc_scores <- rowMeans(progeny_tumour[samples, onc_pathways_found, drop = FALSE])
imi_scores <- rowMeans(progeny_stroma[samples, imi_pathways_found, drop = FALSE])
delta      <- onc_scores - imi_scores
phenotype  <- ifelse(delta > 0, "ONC", "IMI")

# Add sample metadata
sample_meta <- data.frame(
  sample     = samples,
  primary    = sapply(enhanced_list, function(x) unique(colData(x)$sample)),
  stringsAsFactors = FALSE
)

# Retrieve primary + metastasis from SCE colData if stored; otherwise use sample name
get_meta <- function(sce, field) {
  v <- colData(sce)[[field]]
  if (!is.null(v)) unique(v)[1] else NA_character_
}

result <- data.frame(
  sample      = samples,
  ONC_score   = round(onc_scores, 4),
  IMI_score   = round(imi_scores, 4),
  delta       = round(delta, 4),
  phenotype   = phenotype,
  stringsAsFactors = FALSE
)
rownames(result) <- NULL

print(result)
```

# Tumor and Stroma Combined Classification

```{r}
#| echo: false
#| eval: false
# Optional: Calculate PROGENy scores on total pseudobulk (Tumour + Stroma) for comparison

total_mat <- sapply(pseudobulk_list, function(x) {
  vals <- x$total_counts; names(vals) <- x$gene; vals[all_genes]
})
rownames(total_mat) <- all_genes

# PROGENy scores (samples × pathways)
progeny_total_mat <- progeny(
  total_mat,
  scale        = TRUE,
  organism     = "Human",
  top          = 500,
  perm         = 1,
  z_scores     = TRUE
) |> as.data.frame()

# Keep only the 7 selected pathways
pathways_avail <- intersect(ALL_PATHWAYS, colnames(progeny_total_mat))
if (length(pathways_avail) < length(ALL_PATHWAYS)) {
  missing <- setdiff(ALL_PATHWAYS, pathways_avail)
  warning(sprintf("Pathways not found in PROGENy model: %s", paste(missing, collapse=", ")))
}

progeny_total_mat <- progeny_total_mat[, intersect(ALL_PATHWAYS, colnames(progeny_total_mat)), drop=FALSE]

cat("PROGENy complete.\n")
cat(sprintf("  Total pathways: %s\n", paste(colnames(progeny_total_mat), collapse=", ")))

onc_scores_total <- rowMeans(progeny_total_mat[samples, onc_pathways_found, drop = FALSE])
imi_scores_total <- rowMeans(progeny_total_mat[samples, imi_pathways_found, drop = FALSE])
delta_total      <- onc_scores_total - imi_scores_total
phenotype_total  <- ifelse(delta_total > 0, "ONC", "IMI")
result$total_ONC_score <- round(onc_scores_total, 4)
result$total_IMI_score <- round(imi_scores_total, 4)
result$total_delta     <- round(delta_total, 4)
result$total_phenotype <- phenotype_total

```



```{r heatmap, fig.height=7, fig.width=10}
#| echo: false
#| eval: false
## Pathway heatmap

# Combine tumour and stroma scores for heatmap
# Label each row as "<sample> (T)" or "<sample> (S)"
tumour_rows <- progeny_tumour[samples, , drop = FALSE]
stroma_rows <- progeny_stroma[samples, , drop = FALSE]

rownames(tumour_rows) <- paste0(samples, " [T]")
rownames(stroma_rows) <- paste0(samples, " [S]")

# Column order: ONC pathways first, then IMI
col_order   <- intersect(c(PATHWAYS_ONC, PATHWAYS_IMI),
                         union(colnames(tumour_rows), colnames(stroma_rows)))
heatmap_mat <- rbind(
  tumour_rows[, intersect(col_order, colnames(tumour_rows)), drop=FALSE],
  stroma_rows[, intersect(col_order, colnames(stroma_rows)), drop=FALSE]
)

# Annotation: ONC / IMI pathway group + Tumour / Stroma compartment
col_annot <- data.frame(
  pathway_group = ifelse(col_order %in% PATHWAYS_ONC, "ONC", "IMI"),
  row.names     = col_order
)
row_annot <- data.frame(
  compartment = rep(c("Tumour", "Stroma"), each = length(samples)),
  phenotype   = rep(result$phenotype, 2),
  row.names   = rownames(heatmap_mat)
)

pheatmap(heatmap_mat,
         annotation_col  = col_annot,
         annotation_row  = row_annot,
         cluster_cols    = FALSE,
         cluster_rows    = TRUE,
         scale           = "column",
         color           = colorRampPalette(c("#2166AC","white","#D6604D"))(100),
         fontsize        = 10,
         main            = "PROGENy pathway scores — Tumour [T] and Stroma [S] pseudobulks")
```

## ONC / IMI scatter plot

```{r scatter, fig.height=5, fig.width=7}
ggplot(result, aes(ONC_score, IMI_score, colour = phenotype, label = sample)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey40") +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_colour_manual(values = c("ONC" = "#E07B54", "IMI" = "#4393C3")) +
  labs(title = "PROGENy ONC vs IMI scores - Visium metastatic samples",
       subtitle = "Points above diagonal = IMI; below = ONC",
       x = "ONC score (mean PI3K, EGFR, MAPK, VEGF - Tumour)",
       y = "IMI score (mean JAK-STAT, NFkB, TNFa - Stroma)",
       colour = "Phenotype") +
  theme_classic(base_size = 12)
```

## Save results

```{r save}
write.csv(result, OUT_CSV, row.names = FALSE)
cat("Saved → ", OUT_CSV, "\n")
print(table(result$phenotype))
```

## Session info

```{r session}
sessionInfo()
```
