---
title: "V06 — Tumour Niche Definition"
author: "Abu Saadat"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: "right"
    toc-title: "Contents"
    toc-expand: true
    code-fold: true
    code-tools: false
    code-summary: "Hide code"
    lightbox: true
    self-contained: true
    fig-width: 12
    fig-height: 7
    fig-dpi: 150
    cap-location: bottom
    fontsize: 1.05em
    linestretch: 1.6
    theme:
      light: "cosmo"
      dark : "cyborg"
execute:
  warning: false
  message: false
  echo: false
---

Classifies each enhanced sub-spot into a spatial niche zone based on physical distance to
the nearest Tumour sub-spot:

| Zone | Threshold |
|------|-----------|
| **Tumour** | sub-spot is Tumour (compartment == "Tumour") |
| **Peritumoral** | non-Tumour, distance < 100 µm |
| **Distal** | distance ≥ 100 µm |

Note: the Xenium pipeline uses a finer "Tumour niche" zone (< 30 µm) that is not applicable
here — BayesSpace sub-spots with `nsubspots.per.edge = 3` have a minimum inter-sub-spot
spacing of ~33 µm (100 µm original pitch ÷ 3), so the 30 µm zone is sub-resolution for Visium
and has been merged into Peritumoral.

**Distance calculation**: pixel coordinates (`pxl_col_in_fullres`, `pxl_row_in_fullres`) from
the Visium spatial directory are converted to µm using a per-sample calibration factor:
`microns_per_pixel = 55 / spot_diameter_fullres`, where `spot_diameter_fullres` (spot diameter
in full-resolution image pixels) is read from each sample's `scalefactors_json.json`.
Array coordinates are **not** used: the Visium hex grid has `array_col` incrementing by 2
between same-row neighbours and uses non-isotropic row/column scales, making array-unit
distances incorrect. Pixel coordinates are isotropic and give true Euclidean distances.
Distances are computed with `RANN::nn2()`.

**Input**: `visium_annotated_sce_list.rds`
**Output**: `visium_niches_sce_list.rds`

```{r setup}
library(SingleCellExperiment)
library(RANN)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)

BASE    <- "."
OUT_RDS <- file.path(BASE, "visium_niches_ucd_sce_list.rds")

# Niche distance threshold (µm)
# "Tumour niche" (< 30 µm) collapsed into Peritumoral: BayesSpace sub-spot spacing
# (~33 µm with nsubspots.per.edge=3) is coarser than the 30 µm Xenium threshold.
NICHE_OUTER      <- 100   # Peritumoral: < 100 µm; Distal: ≥ 100 µm

# Known Visium spot diameter (capture area) = 55 µm
# Used with spot_diameter_fullres (pixels) from scalefactors_json.json
# to derive a per-sample microns-per-pixel conversion factor.
SPOT_DIAMETER_UM <- 55

enhanced_list <- readRDS(file.path(BASE, "visium_celltypes_ucd_sce_list.rds"))
# sample_dirs derived dynamically — only scalefactors_json.json is needed
sample_dirs <- setNames(file.path(BASE, names(enhanced_list)), names(enhanced_list))
```

## Assign niche zones per sample

```{r assign-niches, results='asis'}
for (s in names(enhanced_list)) {
  cat(sprintf("\n---\n\n### %s\n\n", s))
  sce_enh <- enhanced_list[[s]]

  cd         <- as.data.frame(colData(sce_enh))
  is_tumour  <- cd$compartment == "Tumour"
  n_tumour   <- sum(is_tumour, na.rm = TRUE)
  n_nontumour <- sum(!is_tumour, na.rm = TRUE)

  cat(sprintf("  %d tumour sub-spots, %d non-tumour sub-spots\n\n",
              n_tumour, n_nontumour))

  if (n_tumour == 0) {
    warning(sprintf("%s: no Tumour sub-spots — all sub-spots set to Distal", s))
    colData(sce_enh)$niche <- "Distal"
    enhanced_list[[s]] <- sce_enh
    next
  }

  # Convert pixel coordinates to physical µm via per-sample calibration
  sf_path <- file.path(sample_dirs[s], "spatial", "scalefactors_json.json")
  sf      <- jsonlite::fromJSON(sf_path)
  mpp     <- SPOT_DIAMETER_UM / sf$spot_diameter_fullres   # microns per pixel
  cat(sprintf("  spot_diameter_fullres = %.2f px → %.4f µm/px\n\n",
              sf$spot_diameter_fullres, mpp))

  # Use parent coordinates stored in colData by V02 (from Seurat image @coordinates).
  # parent_array_row/col and pxl_row/col_in_fullres identify the physical parent spot;
  # scale is derived by linear regression across all parent spots in this sample.
  parent_df    <- unique(data.frame(
    array_row = cd$parent_array_row,
    array_col = cd$parent_array_col,
    pxl_row   = cd$pxl_row_in_fullres,
    pxl_col   = cd$pxl_col_in_fullres
  ))
  scale_col_px <- coef(lm(pxl_col ~ array_col, data = parent_df))[["array_col"]]
  scale_row_px <- coef(lm(pxl_row ~ array_row, data = parent_df))[["array_row"]]
  cat(sprintf("  scale: %.1f px/col-unit (%.1f µm), %.1f px/row-unit (%.1f µm)\n\n",
              scale_col_px, scale_col_px * mpp,
              scale_row_px, scale_row_px * mpp))

  parent_pxl_col <- cd$pxl_col_in_fullres
  parent_pxl_row <- cd$pxl_row_in_fullres
  parent_arr_col <- cd$parent_array_col
  parent_arr_row <- cd$parent_array_row

  # Inferred sub-spot pixel position = physical parent centre + array offset × scale
  coords_um <- data.frame(
    x = (parent_pxl_col + (cd$array_col - parent_arr_col) * scale_col_px) * mpp,
    y = (parent_pxl_row + (cd$array_row - parent_arr_row) * scale_row_px) * mpp
  )

  # Valid masks — drop sub-spots with NA pixel coords (off-by-one in V02 spot.idx;
  # affects ~1 parent spot per sample; those are assigned Distal)
  valid          <- complete.cases(coords_um)
  is_tumour_v    <- is_tumour  & valid
  is_nontumour_v <- !is_tumour & valid

  if (any(!valid))
    cat(sprintf("  Dropping %d sub-spots with NA pixel coords → Distal\n\n", sum(!valid)))

  tumour_coords    <- coords_um[is_tumour_v,    , drop = FALSE]
  nontumour_coords <- coords_um[is_nontumour_v, , drop = FALSE]

  # Nearest tumour sub-spot for each valid non-tumour sub-spot
  nn_result      <- RANN::nn2(data = tumour_coords, query = nontumour_coords, k = 1)
  dist_to_tumour <- nn_result$nn.dists[, 1]  # µm

  # Classify valid non-tumour sub-spots
  niche_nontumour_v <- dplyr::case_when(
    dist_to_tumour <  NICHE_OUTER ~ "Peritumoral",
    TRUE                          ~ "Distal"
  )

  # Combine: valid Tumour | valid non-tumour classified | NA-coord → Distal
  niche_all <- character(nrow(cd))
  niche_all[is_tumour_v]    <- "Tumour"
  niche_all[is_nontumour_v] <- niche_nontumour_v
  niche_all[!valid]         <- "Distal"

  colData(sce_enh)$niche          <- niche_all
  colData(sce_enh)$dist_to_tumour <- NA_real_
  colData(sce_enh)$dist_to_tumour[is_nontumour_v] <- dist_to_tumour

  enhanced_list[[s]] <- sce_enh

  # Report
  niche_fac    <- factor(niche_all, levels = c("Tumour", "Peritumoral", "Distal"))
  niche_counts <- table(niche_fac)
  niche_pct    <- round(100 * niche_counts / sum(niche_counts), 1)
  for (zone in c("Tumour", "Peritumoral", "Distal")) {
    cat(sprintf("  %-20s %5d sub-spots (%s%%)\n",
                zone,
                niche_counts[zone],
                niche_pct[zone]))
  }
  cat("\n")
}
```

## Spatial niche maps

```{r niche-maps, fig.height=4.5, results='asis'}
niche_colours <- c(
  "Tumour"      = "#C0392B",
  "Peritumoral" = "#E67E22",
  "Distal"      = "#BDC3C7"
)
niche_levels <- c("Tumour", "Peritumoral", "Distal")

for (s in names(enhanced_list)) {
  cat(sprintf("\n#### %s\n\n", s))
  sce_enh <- enhanced_list[[s]]

  df <- data.frame(
    col   = colData(sce_enh)$array_col,
    row   = colData(sce_enh)$array_row,
    niche = factor(colData(sce_enh)$niche, levels = niche_levels)
  )

  p <- ggplot(df, aes(col, -row, colour = niche)) +
    geom_point(size = 0.25, alpha = 0.8) +
    coord_fixed() +
    scale_colour_manual(values = niche_colours, drop = FALSE) +
    labs(title = s, colour = NULL) +
    theme_classic(base_size = 11) +
    theme(
      panel.background = element_rect(fill = "black"),
      plot.background  = element_rect(fill = "black"),
      axis.line        = element_line(colour = "white"),
      axis.text        = element_text(colour = "white"),
      axis.title       = element_text(colour = "white"),
      plot.title       = element_text(colour = "white"),
      legend.background = element_rect(fill = "black"),
      legend.text      = element_text(colour = "white"),
      legend.key       = element_rect(fill = "black")
    ) +
    guides(colour = guide_legend(override.aes = list(size = 2.5)))

  print(p)
  cat("\n")
}

# grid all plots together
# (note: this is a bit hacky since the individual plots are printed in a loop, but it works)
 plot_list <- lapply(names(enhanced_list), function(s) {
   sce_enh <- enhanced_list[[s]]
   df <- data.frame(
     col   = colData(sce_enh)$array_col,
     row   = colData(sce_enh)$array_row,
     niche = factor(colData(sce_enh)$niche, levels = niche_levels)
   )
   ggplot(df, aes(col, -row, colour = niche)) +
     geom_point(size = 0.25, alpha = 0.8) +
     coord_fixed() +
     scale_colour_manual(values = niche_colours, drop = FALSE) +
     labs(title = s, colour = NULL) +
     theme_classic(base_size = 11) +
     theme(
       panel.background = element_rect(fill = "black"),
       plot.background  = element_rect(fill = "black"),
       axis.line        = element_line(colour = "white"),
       axis.text        = element_text(colour = "white"),
       axis.title       = element_text(colour = "white"),
       plot.title       = element_text(colour = "white"),
       legend.background = element_rect(fill = "black"),
       legend.text      = element_text(colour = "white"),
       legend.key       = element_rect(fill = "black")
     ) +
     guides(colour = guide_legend(override.aes = list(size = 2.5)))
 })
 combined_plot <- wrap_plots(plot_list, ncol = 4) +
   plot_annotation(title = "Spatial niche maps", theme = theme(plot.title = element_text(size = 16, face = "bold")))
 print(combined_plot)

# Save the combined plot as a high-resolution image
 ggsave(filename = file.path(BASE, "spatial_niche_maps.png"), plot = combined_plot, width = 16, height = 12, dpi = 300)

```

## Distance distribution plots

```{r dist-plots, fig.height=4}
dist_df <- lapply(names(enhanced_list), function(s) {
  cd <- as.data.frame(colData(enhanced_list[[s]]))
  cd_nontumour <- cd[cd$compartment == "Stroma" & !is.na(cd$dist_to_tumour), ]
  data.frame(sample = s, dist_um = cd_nontumour$dist_to_tumour)
}) |> bind_rows()

ggplot(dist_df, aes(dist_um, colour = sample)) +
  geom_density(linewidth = 0.7) +
  geom_vline(xintercept = NICHE_OUTER,
             linetype = "dashed", colour = "grey40") +
  annotate("text", x = NICHE_OUTER, y = Inf, label = "100 µm",
           hjust = -0.1, vjust = 1.5, size = 3) +
  scale_x_log10() +
  labs(title = "Distance of Stroma sub-spots to nearest Tumour sub-spot",
       x = "Distance to nearest tumour sub-spot (µm, log scale)",
       y = "Density", colour = "Sample") +
  theme_classic(base_size = 11)
```

## Niche composition summary

```{r summary-table}
summary_df <- lapply(names(enhanced_list), function(s) {
  niche <- colData(enhanced_list[[s]])$niche
  n     <- length(niche)
  data.frame(
    sample           = s,
    n_subspots       = n,
    pct_tumour       = round(100 * mean(niche == "Tumour"),       1),
    pct_peritumoral  = round(100 * mean(niche == "Peritumoral"),  1),
    pct_distal       = round(100 * mean(niche == "Distal"),       1)
  )
}) |> bind_rows()

knitr::kable(summary_df, caption = "Niche zone composition per sample (% of enhanced sub-spots)")
```

## Immune cell composition by niche zone and ONC/IMI phenotype

Immune cell type counts in Peritumoral and Distal zones, grouped by the
ONC/IMI classification from V05 (`metadata/ONC_IMI_phenotypes.csv`).
Tumour sub-spots are excluded (they carry the "Tumour" label, not an immune type).

```{r immune-niche-barplot, fig.width = 14, fig.height = 6}
marta_csv <- file.path(BASE, "metadata", "ONC_IMI_phenotypes.csv")

  pheno_df  <- read.csv(marta_csv, stringsAsFactors = FALSE)
  pheno_map <- setNames(pheno_df[,2], sub("^CUP-", "", pheno_df[,1]))

  immune_df <- lapply(names(enhanced_list), function(s) {
    cd    <- as.data.frame(colData(enhanced_list[[s]]))
    pheno <- pheno_map[s]
    if (is.na(pheno)) return(NULL)
    cd |>
      filter(niche %in% c("Peritumoral", "Distal"),
             celltype != "Stroma") |>
      mutate(
        sample    = s,
        phenotype = pheno,
        celltype  = gsub("_", " ", celltype)
      ) |>
      count(sample, phenotype, niche, celltype, name = "n_subspots")
  }) |> bind_rows()

# Aggregate counts across samples within each phenotype group
    plot_df <- immune_df |>
      group_by(phenotype, niche, celltype) |>
      summarise(total = sum(n_subspots), .groups = "drop") |>
      mutate(
        niche     = factor(niche, levels = c("Peritumoral", "Distal")),
        phenotype = factor(phenotype, levels = c("ONC", "IMI"))
      )

# Stacked barplot Celltype distribution in Niche+Peritumoral vs Distal

plot_df <- immune_df %>%
  group_by(celltype) %>%
  mutate(total_n = sum(n_subspots),
  Location = case_when(
        niche %in% c("Niche", "Peritumoral") ~ "Niche/Peritumoral",
        niche == "Distal" ~ "Distal",
        TRUE ~ niche
    )) %>%
  ungroup()

# Barplot of grouped cell_type composition by Location

ggplot(plot_df, aes(x = reorder(celltype, -total_n), y = n_subspots, fill = Location)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_manual(values = c("Niche/Peritumoral" = "#0c797f", "Distal" = "#b45221")) +
  labs(
    title = "Cell Type Counts by Location (CUP)",
    x = "",
    y = "Sub-spot Count",
    fill = "Location"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 11),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

dir.create(file.path(BASE, "Plots"), showWarnings = FALSE)
ggsave(file.path(BASE, "Plots", "cell_type_counts_by_location_CUP.png"), width = 10, height = 6, dpi = 300)

ggplot(plot_df, aes(x = Location, y = n_subspots, fill = Location)) +
  geom_boxplot() +
  facet_wrap(~ celltype, scales = "free_y") +
  scale_fill_manual(values = c("Niche/Peritumoral" = "#0c797f", "Distal" = "#b45221")) +
  labs(
    title = "Cell Type Counts by Location (CUP)",
    x = "",
    y = "Sub-spot Count",
    fill = "Location"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 11),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

ggsave(file.path(BASE, "Plots", "cell_type_counts_by_location_boxplot_CUP.png"), width = 12, height = 8, dpi = 300)
```

## Cell type counts by ONC/IMI phenotype (all zones combined)

Per-sample sub-spot counts for each immune cell type, collapsed across Peritumoral and
Distal zones. One point per sample; ONC vs IMI compared with a two-sided
Wilcoxon rank-sum test (p-value shown above each panel).

```{r celltype-counts-onc-imi, fig.width = 12, fig.height = 6}
marta_csv <- file.path(BASE, "metadata", "ONC_IMI_phenotypes.csv")

if (file.exists(marta_csv)) {
  pheno_df  <- read.csv(marta_csv, stringsAsFactors = FALSE)
  pheno_map <- setNames(pheno_df[,2], sub("^CUP-", "", pheno_df[,1]))

  counts_onc_imi <- lapply(names(enhanced_list), function(s) {
    cd    <- as.data.frame(colData(enhanced_list[[s]]))
    pheno <- pheno_map[s]
    if (is.na(pheno)) return(NULL)
    cd |>
      filter(niche %in% c("Peritumoral", "Distal"),
             !celltype %in% c("Tumour", "Stroma")) |>
      mutate(sample    = s,
             phenotype = pheno,
             celltype  = gsub("_", " ", celltype)) |>
      count(sample, phenotype, celltype, name = "n_subspots")
  }) |> bind_rows() |>
    mutate(phenotype = factor(phenotype, levels = c("ONC", "IMI")))

  # Wilcoxon p-value and label position per cell type
  ct_stats_counts <- counts_onc_imi |>
    group_by(celltype) |>
    summarise(
      p_value = tryCatch(
        wilcox.test(n_subspots[phenotype == "ONC"],
                    n_subspots[phenotype == "IMI"],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_
      ),
      y_pos = max(n_subspots, na.rm = TRUE) * 1.08,
      .groups = "drop"
    ) |>
    mutate(p_label = dplyr::case_when(
      is.na(p_value)  ~ "NA",
      p_value < 0.001 ~ "p<0.001",
      TRUE            ~ sprintf("p=%.3f", p_value)
    ))

  ct_order        <- ct_stats_counts |> arrange(p_value) |> pull(celltype)
  counts_onc_imi  <- counts_onc_imi  |> mutate(celltype = factor(celltype, levels = ct_order))
  ct_stats_counts <- ct_stats_counts |> mutate(celltype = factor(celltype, levels = ct_order))

  ggplot(counts_onc_imi, aes(x = phenotype, y = n_subspots, fill = phenotype)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
    geom_jitter(aes(colour = phenotype), width = 0.15, size = 2.2, alpha = 0.85) +
    geom_text(data        = ct_stats_counts,
              aes(x = 1.5, y = y_pos, label = p_label),
              inherit.aes = FALSE, size = 3, colour = "grey30") +
    facet_wrap(~ celltype, scales = "free_y", ncol = 4) +
    scale_fill_manual(values  = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                      name = "Phenotype") +
    scale_colour_manual(values = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                        name = "Phenotype") +
    labs(
      title    = "Immune cell type sub-spot counts by ONC/IMI phenotype",
      x        = NULL,
      y        = "Sub-spot count"
    ) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1, size = 9),
      legend.position = "bottom",
      plot.title      = element_text(face = "bold"),
      strip.text      = element_text(size = 9, face = "bold")
    )

  ggsave(file.path(BASE, "Plots", "celltype_counts_onc_imi.png"),
         width = 12, height = 6, dpi = 300)
} else {
  message("ONC_IMI_phenotypes.csv not found — skipping ONC/IMI cell type count plot.")
}
```

## Immune cell log₂ enrichment by niche zone and ONC/IMI phenotype

Mirrors the `log2_enrich` calculation in Xenium Script 14.
For each sample, the global baseline is the proportional composition of all
non-Tumour sub-spots (Peritumoral + Distal combined).

$$\text{log}_2\text{-enrich} = \log_2\!\left(\frac{\%_\text{zone} + 0.001}{\%_\text{global} + 0.001}\right)$$

Positive = cell type enriched in that zone relative to its overall abundance; negative = depleted.

```{r enrich-compute}
marta_csv <- file.path(BASE, "metadata", "ONC_IMI_phenotypes.csv")

enrich_all <- NULL

if (file.exists(marta_csv)) {
  pheno_df  <- read.csv(marta_csv, stringsAsFactors = FALSE)
  pheno_map <- setNames(pheno_df[,2], sub("^CUP-", "", pheno_df[,1]))

  enrich_all <- lapply(names(enhanced_list), function(s) {
    cd    <- as.data.frame(colData(enhanced_list[[s]]))
    pheno <- pheno_map[s]
    if (is.na(pheno)) return(NULL)

    # Non-Tumour sub-spots only (exclude Stroma catch-all if present)
    nt <- cd |> filter(niche %in% c("Peritumoral", "Distal"),
                       !celltype %in% c("Tumour", "Stroma"))

    # Global baseline: proportions across Peritumoral + Distal combined
    global_tab <- table(nt$celltype)
    global_n   <- max(sum(global_tab), 1)

    lapply(c("Peritumoral", "Distal"), function(z) {
      sub_z  <- nt |> filter(niche == z)
      zone_tab <- table(sub_z$celltype)
      zone_n   <- max(sum(zone_tab), 1)
      types    <- union(names(global_tab), names(zone_tab))

      pct_zone   <- as.integer(zone_tab[types])   / zone_n   * 100
      pct_global <- as.integer(global_tab[types]) / global_n * 100
      pct_zone[is.na(pct_zone)]     <- 0
      pct_global[is.na(pct_global)] <- 0

      tibble(
        sample      = s,
        phenotype   = pheno,
        niche       = z,
        cell_type   = gsub("_", " ", types),
        pct         = pct_zone,
        pct_global  = pct_global,
        log2_enrich = log2((pct_zone / 100 + 0.001) / (pct_global / 100 + 0.001))
      )
    }) |> bind_rows()
  }) |> bind_rows()
} else {
  message("V05 classification not found — skipping enrichment plots.")
}
```



```{r enrich-heatmap, fig.width = 13, fig.height = 5}
if (!is.null(enrich_all) && nrow(enrich_all) > 0 &&
    requireNamespace("pheatmap", quietly = TRUE)) {
  library(pheatmap)

  peri_wide <- enrich_all |>
    filter(niche == "Peritumoral") |>
    select(sample, cell_type, log2_enrich) |>
    pivot_wider(names_from = sample, values_from = log2_enrich, values_fill = 0) |>
    column_to_rownames("cell_type") |>
    as.matrix()

  # Column annotation: ONC/IMI phenotype
  tmp_df <- read.csv(file.path(BASE, "metadata", "ONC_IMI_phenotypes.csv"), stringsAsFactors = FALSE)
  pheno_map2 <- setNames(tmp_df[,2], sub("^CUP-", "", tmp_df[,1]))
  col_ann <- data.frame(
    Phenotype = ifelse(is.na(pheno_map2[colnames(peri_wide)]),
                       "Unknown", pheno_map2[colnames(peri_wide)]),
    row.names = colnames(peri_wide)
  )

  max_abs <- max(abs(peri_wide), 2, na.rm = TRUE)
  pal     <- colorRampPalette(
    c("#313695","#4575B4","#74ADD1","#ABD9E9","white",
      "#FDAE61","#F46D43","#D73027","#A50026")
  )(101)
  breaks  <- seq(-max_abs, max_abs, length.out = 102)

  pheatmap(
    peri_wide,
    color             = pal,
    breaks            = breaks,
    cluster_rows      = TRUE,
    cluster_cols      = TRUE,
    annotation_col    = col_ann,
    annotation_colors = list(
      Phenotype = c("ONC" = "#D73027", "IMI" = "#4575B4", "Unknown" = "#888888")
    ),
    border_color      = NA,
    fontsize          = 10,
    fontsize_row      = 9,
    fontsize_col      = 9,
    main              = "Peritumoral immune enrichment (log₂) per sample",
    legend_breaks     = c(floor(-max_abs), 0, ceiling(max_abs)),
    legend_labels     = c(sprintf("Depleted\n(%.0f)", floor(-max_abs)),
                          "Equal\n(0)",
                          sprintf("Enriched\n(%.0f)", ceiling(max_abs)))
  )
}
```

## Spatial immune cell enrichment by ONC/IMI phenotype

Mirrors Xenium Script 17b. For each niche zone and immune cell type, a two-sided
Wilcoxon rank-sum test (ONC vs IMI) is run on per-sample log₂ enrichment
scores with rank-biserial correlation as the effect size. Cell types are ordered by
their maximum |r| across both zones. Zones shown: Peritumoral (< 100 µm) and Distal
(≥ 100 µm), mirroring Script 17b's Niche/Peritumoral and Distal.

```{r spatial-immune-enrichment, fig.width = 14, fig.height = 7}
if (!is.null(enrich_all) && nrow(enrich_all) > 0) {

  # Build wide matrices: samples × cell types, separately per zone
  make_wide_mat <- function(df, zone_name) {
    df |> filter(niche == zone_name) |>
      select(sample, cell_type, log2_enrich) |>
      pivot_wider(names_from = cell_type, values_from = log2_enrich, values_fill = 0) |>
      column_to_rownames("sample") |> as.matrix()
  }

  peri_mat_e   <- make_wide_mat(enrich_all, "Peritumoral")
  distal_mat_e <- make_wide_mat(enrich_all, "Distal")

  # Phenotype vector from the manual phenotype CSV
  pheno_csv_e  <- read.csv(file.path(BASE, "metadata", "ONC_IMI_phenotypes.csv"), stringsAsFactors = FALSE)
  pheno_vec_e  <- setNames(pheno_csv_e[,2], sub("^CUP-", "", pheno_csv_e[,1]))
  common_e     <- intersect(rownames(peri_mat_e), names(pheno_vec_e[!is.na(pheno_vec_e)]))
  pheno_vec_e  <- pheno_vec_e[common_e]

  # Effect size helper: rank-biserial r (matches Script 17b)
  rank_bis_r <- function(x, y) {
    n1 <- sum(!is.na(x)); n2 <- sum(!is.na(y))
    if (n1 < 2 || n2 < 2) return(NA_real_)
    W <- suppressWarnings(wilcox.test(x, y, exact = FALSE)$statistic)
    as.numeric(1 - (2 * W) / (n1 * n2))
  }

  compute_zone_stats <- function(mat, pheno_vec, zone_name) {
    samps <- intersect(rownames(mat), names(pheno_vec))
    bind_rows(lapply(colnames(mat), function(ct) {
      onc <- mat[samps[pheno_vec[samps] == "ONC"], ct, drop = TRUE]
      imi <- mat[samps[pheno_vec[samps] == "IMI"], ct, drop = TRUE]
      p   <- suppressWarnings(tryCatch(
        wilcox.test(onc, imi, exact = FALSE)$p.value,
        error = function(e) NA_real_))
      tibble(cell_type = ct, niche = zone_name,
             rank_biserial = rank_bis_r(onc, imi), p_value = p)
    }))
  }

  stats_e <- bind_rows(
    compute_zone_stats(peri_mat_e,   pheno_vec_e, "Peritumoral"),
    compute_zone_stats(distal_mat_e, pheno_vec_e, "Distal")
  )

  # Order cell types by maximum |r| across both zones
  ct_order_e <- stats_e |>
    group_by(cell_type) |>
    summarise(max_r = max(abs(rank_biserial), na.rm = TRUE), .groups = "drop") |>
    arrange(desc(max_r)) |> pull(cell_type)

  n_onc_e <- sum(pheno_vec_e == "ONC")
  n_imi_e <- sum(pheno_vec_e == "IMI")

  # Panel labels: cell_type × zone, Peritumoral panels first
  panel_levels_e <- c(paste0(ct_order_e, "\n[Peritumoral]"),
                      paste0(ct_order_e, "\n[Distal]"))

  long_e <- enrich_all |>
    filter(sample %in% common_e) |>
    mutate(
      panel_label = factor(paste0(cell_type, "\n[", niche, "]"),
                           levels = panel_levels_e),
      phenotype   = factor(phenotype, levels = c("ONC", "IMI"))
    )

  stats_e <- stats_e |>
    mutate(
      panel_label = factor(paste0(cell_type, "\n[", niche, "]"),
                           levels = panel_levels_e),
      eff_label   = sprintf("r = %.2f\np = %.3f", rank_biserial, p_value)
    )

  caption_e <- paste0(
    "Log₂ enrichment of immune cell types relative to global tissue fraction, ",
    "stratified by ONC (red, n=", n_onc_e, ") and IMI (blue, n=", n_imi_e, "). ",
    "Left: Peritumoral zone (sub-spots < 100 µm from tumour). ",
    "Right: Distal zone (≥ 100 µm). ",
    "Columns ordered by maximum |rank-biserial r| across zones. ",
    "r = rank-biserial correlation (effect size); ",
    "p = two-sided Wilcoxon rank-sum test (unadjusted)."
  )

  ggplot(long_e, aes(x = phenotype, y = log2_enrich, fill = phenotype)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
    geom_jitter(aes(colour = phenotype), width = 0.12, size = 2.2, alpha = 0.85) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
    geom_text(data        = stats_e,
              aes(x = 1.5, y = Inf, label = eff_label),
              vjust = 1.4, size = 3.2, colour = "grey40", inherit.aes = FALSE) +
    facet_wrap(~ panel_label, ncol = length(ct_order_e), scales = "free_y") +
    scale_fill_manual(values  = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                      name = "Phenotype") +
    scale_colour_manual(values = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                        name = "Phenotype") +
    labs(
      title   = "Spatial immune cell enrichment by ONC/IMI phenotype",
      x       = NULL,
      y       = expression(log[2]~"enrichment vs global fraction"),
      caption = caption_e
    ) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x           = element_text(angle = 30, hjust = 1, size = 9),
      legend.position       = "bottom",
      plot.title            = element_text(face = "bold"),
      strip.text            = element_text(size = 8, face = "bold"),
      plot.caption          = element_text(size = 7.5, colour = "grey30",
                                           hjust = 0, lineheight = 1.3,
                                           margin = margin(t = 8)),
      plot.caption.position = "plot"
    )
} else {
  message("Enrichment data not available — run enrich-compute chunk first.")
}
```

```{r focused-immune-enrichment, fig.width = 8, fig.height = 7}
if (!is.null(enrich_all) && nrow(enrich_all) > 0 &&
    exists("stats_e") && exists("long_e")) {

  focus_cts <- c("Dendritic", "B_cells", "Monocytes")

  panel_levels_f <- c(paste0(focus_cts, "\n[Peritumoral]"),
                      paste0(focus_cts, "\n[Distal]"))

  long_f <- long_e |>
    filter(cell_type %in% focus_cts) |>
    mutate(panel_label = factor(paste0(cell_type, "\n[", niche, "]"),
                                levels = panel_levels_f))

  stats_f <- stats_e |>
    filter(cell_type %in% focus_cts) |>
    mutate(panel_label = factor(paste0(cell_type, "\n[", niche, "]"),
                                levels = panel_levels_f))

  ggplot(long_f, aes(x = phenotype, y = log2_enrich, fill = phenotype)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
    geom_jitter(aes(colour = phenotype), width = 0.12, size = 3, alpha = 0.85) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
    geom_text(data        = stats_f,
              aes(x = 1.5, y = Inf, label = eff_label),
              vjust = 1.4, size = 3.5, colour = "grey40", inherit.aes = FALSE) +
    facet_wrap(~ panel_label, ncol = 2, scales = "free_y") +
    scale_fill_manual(values  = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                      name = "Phenotype") +
    scale_colour_manual(values = c("ONC" = "#D73027", "IMI" = "#4575B4"),
                        name = "Phenotype") +
    labs(
      title    = "Spatial immune cell enrichment by ONC/IMI phenotype",
      subtitle = "Peritumoral (< 100 µm) and Distal (≥ 100 µm) zones",
      x        = NULL,
      y        = expression(log[2]~"enrichment vs global fraction")
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1, size = 10),
      legend.position = "bottom",
      plot.title      = element_text(face = "bold"),
      strip.text      = element_text(size = 10, face = "bold")
    )
}

ggsave(file.path(BASE, "Plots", "immune_enrichment_by_phenotype.png"),
       width = 8, height = 10, dpi = 300)
```

## Data export

```{r data-export}
library(openxlsx)
wb <- createWorkbook()

# Sheet 1: per-sample cell type sub-spot counts (data behind celltype_counts_onc_imi.png)
if (exists("counts_onc_imi") && nrow(counts_onc_imi) > 0) {
  addWorksheet(wb, "celltype_counts")
  writeData(wb, "celltype_counts",
            counts_onc_imi |>
              arrange(celltype, phenotype, sample) |>
              select(sample, phenotype, celltype, n_subspots))
}

# Sheet 2: Wilcoxon stats for cell type counts ONC vs IMI
if (exists("ct_stats_counts") && nrow(ct_stats_counts) > 0) {
  addWorksheet(wb, "celltype_counts_stats")
  writeData(wb, "celltype_counts_stats",
            ct_stats_counts |>
              arrange(p_value) |>
              select(celltype, p_value, p_label))
}

# Sheet 3: per-sample log2 enrichment scores (data behind spatial enrichment plots)
if (!is.null(enrich_all) && nrow(enrich_all) > 0) {
  addWorksheet(wb, "log2_enrichment")
  writeData(wb, "log2_enrichment",
            enrich_all |>
              arrange(cell_type, niche, phenotype, sample) |>
              select(sample, phenotype, niche, cell_type, pct, pct_global, log2_enrich))
}

# Sheet 4: Wilcoxon stats for log2 enrichment ONC vs IMI (per cell type × niche)
if (exists("stats_e") && nrow(stats_e) > 0) {
  addWorksheet(wb, "enrichment_stats")
  writeData(wb, "enrichment_stats",
            stats_e |>
              arrange(niche, p_value) |>
              select(cell_type, niche, rank_biserial, p_value))
}

out_xlsx <- file.path(BASE, "Plots", "V06_plot_data.xlsx")
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat(sprintf("Exported plot data → %s\n", out_xlsx))
cat(sprintf("  celltype_counts       : %d rows\n",
            if (exists("counts_onc_imi"))  nrow(counts_onc_imi)  else 0))
cat(sprintf("  celltype_counts_stats : %d rows\n",
            if (exists("ct_stats_counts")) nrow(ct_stats_counts) else 0))
cat(sprintf("  log2_enrichment       : %d rows\n",
            if (!is.null(enrich_all))      nrow(enrich_all)      else 0))
cat(sprintf("  enrichment_stats      : %d rows\n",
            if (exists("stats_e"))         nrow(stats_e)         else 0))
```

## Save checkpoint

```{r save}
saveRDS(enhanced_list, OUT_RDS)
cat(sprintf("Saved %d SCEs with niche labels → %s\n", length(enhanced_list), OUT_RDS))
```

## Session info

```{r session}
sessionInfo()
```
