---
title: "Xenium CUP/IC — Cohort-Level Marker Analysis"
subtitle: "Actionable targets & immune checkpoints analysis"
author: "Abu Saadat"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    self-contained: true
    theme: cosmo
    fig-width: 12
    fig-height: 7
    fig-dpi: 150
    cap-location: bottom
execute:
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center")
```

# Configuration

```{r config}

# ── Paths ──────────────────────────────────────────────────────────────────────
ANALYSIS_ROOT <- "."
COHORT_RDS  <- file.path(ANALYSIS_ROOT, "cohort_slim_v2.rds")
OUT_DIR     <- file.path(ANALYSIS_ROOT, "PLOTs", "03_classification")

dir.create(file.path(OUT_DIR, "actionable_targets"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "ICI"),                recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_DIR, "tables"),             recursive = TRUE, showWarnings = FALSE)

# ── Marker gene groups ─────────────────────────────────────────────────────────
GENES_HISTOLOGY  <- "KRT20"
GENES_ACTIONABLE <- c("MET", "ERBB2", "KIT", "FGFR2", "CCNE1", "MYC")
GENES_ICI        <- c("CD274", "LAG3", "CTLA4", "TIGIT", "PDCD1")

# Display labels for axes / titles
GENE_DISPLAY <- c(
  CD274 = "PD-L1 (CD274)",
  PDCD1 = "PD-1 (PDCD1)",
  KRT20 = "CK20 (KRT20)",
  ERBB2 = "ERBB2 (HER2)"
)
label_gene <- function(g) ifelse(g %in% names(GENE_DISPLAY), GENE_DISPLAY[g], g)

# ── Colour scheme ──────────────────────────────────────────────────────────────
COL_TUMOR   <- "#B22222"
COL_STROMA  <- "#4393C3"
COL_COMP    <- c("Tumor" = COL_TUMOR, "Stroma" = COL_STROMA, "Unassigned" = "grey60")

# Sample colour palette (16 distinct colours)
SAMPLE_COLS <- setNames(
  c("#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00","#A65628",
    "#F781BF","#999999","#66C2A5","#FC8D62","#8DA0CB","#E78AC3",
    "#A6D854","#FFD92F","#E5C494","#B3B3B3"),
  character(0)  # will be assigned after loading data
)

```

---

# Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
  library(kableExtra)
  library(ggbeeswarm)   # install.packages("ggbeeswarm") if missing
  library(rstatix)      # install.packages("rstatix") if missing — Wilcoxon p-values
  library(ggpubr)       # install.packages("ggpubr") if missing — stat_compare_means
})
```

---

# Load Cohort Data

```{r load-data}

if (!file.exists(COHORT_RDS)) {
  stop("cohort_slim.rds not found: ", COHORT_RDS,
       "\nRun 01_extract_cohort_data.R first.")
}

cohort <- readRDS(COHORT_RDS)

# Assign consistent sample colours
sample_names <- sort(unique(cohort$sample_name))
SAMPLE_COLS  <- setNames(
  scales::hue_pal()(length(sample_names)),
  sample_names
)

# Filter to Tumor + Stroma only (exclude Unassigned from expression analyses)
cohort_ts <- cohort |>
  filter(Compartment_binary %in% c("Tumor", "Stroma"))

cat(sprintf(
  "Cohort loaded:\n  Samples    : %d\n  Total cells: %d\n  After QC filter (Tumor+Stroma): %d\n",
  length(sample_names), nrow(cohort), nrow(cohort_ts)
))
```

---

# Sample Overview Table

```{r sample-overview-table}

overview <- cohort |>
  group_by(sample_name) |>
  summarise(
    Total_cells    = n(),
    Tumor_cells    = sum(Compartment_binary == "Tumor"),
    Stroma_cells   = sum(Compartment_binary == "Stroma"),
    Unassigned     = sum(Compartment_binary == "Unassigned"),
    Pct_Tumor      = round(100 * Tumor_cells  / Total_cells, 1),
    Pct_Stroma     = round(100 * Stroma_cells / Total_cells, 1),
    Median_CF_score = round(median(CF_score, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  arrange(sample_name)

write.csv(overview,
          file.path(OUT_DIR, "tables", "01_sample_overview.csv"),
          row.names = FALSE)

overview |>
  kable(caption = "Sample overview — cell counts and compartment distribution") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

# Actionable Targets

## Violin plots — expression per sample {#violin-actionable}

*One violin per sample, split by compartment (Tumor = red, Stroma = blue).
Showing only expressing cells (log-normalised expression > 0) to reveal true
signal distribution without zero-inflation collapsing the dynamic range.*

```{r violin-actionable, fig.height=14, fig.width=16}

# Long format — filter to expressing cells only (expression > 0)
# Zero-inflated spatial data collapses violin/boxplots without this filter
action_long <- cohort_ts |>
  dplyr::select(cell_id, sample_name, Compartment_binary, all_of(GENES_ACTIONABLE)) |>
  pivot_longer(cols = all_of(GENES_ACTIONABLE),
               names_to  = "gene",
               values_to = "expression") |>
  filter(!is.na(expression), expression > 0) |>
  mutate(gene = factor(gene, levels = GENES_ACTIONABLE))

p_violin_list <- lapply(GENES_ACTIONABLE, function(g) {
  ggplot(
    action_long |> filter(gene == g),
    aes(x = sample_name, y = expression, fill = Compartment_binary)
  ) +
    geom_violin(
      trim      = TRUE,
      scale     = "width",
      alpha     = 0.75,
      linewidth = 0.3
    ) +
    geom_boxplot(
      width         = 0.15,
      outlier.shape = NA,
      position      = position_dodge(0.9),
      fill          = "white",
      linewidth     = 0.3
    ) +
    scale_fill_manual(values = COL_COMP, name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(
      title = label_gene(g),
      x     = NULL,
      y     = "Log-normalised expression"
    ) +
    theme_classic(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 12),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "top"
    )
})

p_violin_action <- wrap_plots(p_violin_list, ncol = 2)

ggsave(
  file.path(OUT_DIR, "actionable_targets", "violin_actionable_targets.png"),
  p_violin_action,
  width  = 16,
  height = 14,
  dpi    = 300,
  bg     = "white"
)

p_violin_action
```

## Dot plot — mean expression × % expressing cells {#dotplot-actionable}

*Each dot = one sample. Dot size = percentage of tumor cells with detectable
expression (> 0). Dot colour = mean log-normalised expression in tumor cells.
Threshold for "expressing": raw expression > 0 (any detected transcript).*

```{r dotplot-actionable, fig.height=6, fig.width=10}

dot_df <- cohort_ts |>
  filter(Compartment_binary == "Tumor") |>
  dplyr::select(sample_name, all_of(GENES_ACTIONABLE)) |>
  pivot_longer(cols = all_of(GENES_ACTIONABLE),
               names_to  = "gene",
               values_to = "expression") |>
  filter(!is.na(expression)) |>
  group_by(sample_name, gene) |>
  summarise(
    mean_expr   = mean(expression, na.rm = TRUE),
    pct_express = 100 * mean(expression > 0, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  mutate(gene = factor(gene, levels = GENES_ACTIONABLE))

p_dot_action <- ggplot(dot_df,
    aes(x = gene, y = sample_name,
        size  = pct_express,
        color = mean_expr)) +
  geom_point(alpha = 0.85) +
  scale_size_continuous(
    name   = "% expressing\ntumor cells",
    range  = c(1, 12),
    breaks = c(10, 25, 50, 75, 100)
  ) +
  scale_color_gradientn(
    name   = "Mean log-norm\nexpression",
    colors = c("grey90", "#FDBB84", "#E34A33", "#7F0000")
  ) +
  scale_x_discrete(labels = label_gene) +
  labs(
    title    = "Actionable Targets — Tumor Cells",
    subtitle = "Dot size = % expressing cells | Colour = mean expression",
    x        = NULL,
    y        = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    axis.text.x   = element_text(angle = 30, hjust = 1, face = "bold", size = 11),
    axis.text.y   = element_text(size = 9),
    legend.position = "right"
  )

ggsave(
  file.path(OUT_DIR, "actionable_targets", "dotplot_actionable_targets.png"),
  p_dot_action,
  width  = 10,
  height = 6,
  dpi    = 300,
  bg     = "white"
)

p_dot_action
```

## Cell counts expressing each marker {#counts-expressing}

*Number of cells (absolute count) with detectable expression (> 0) per sample,
split by Tumor and Stroma. Gives an immediate sense of which markers have
meaningful signal and in which compartment before looking at expression levels.
Bars are stacked: Tumor on top of Stroma.*

```{r counts-expressing, fig.height=14, fig.width=16}

# All markers — histology + actionable + ICI
ALL_GENES_PLOT <- c(GENES_HISTOLOGY, GENES_ACTIONABLE, GENES_ICI)

# Count expressing cells per sample x gene x compartment
counts_expr <- cohort_ts |>
  dplyr::select(cell_id, sample_name, Compartment_binary,
                all_of(ALL_GENES_PLOT)) |>
  pivot_longer(cols      = all_of(ALL_GENES_PLOT),
               names_to  = "gene",
               values_to = "expression") |>
  filter(!is.na(expression)) |>
  group_by(sample_name, gene, Compartment_binary) |>
  summarise(
    n_expressing = sum(expression > 0, na.rm = TRUE),
    n_total      = n(),
    pct_pos      = 100 * n_expressing / n_total,
    .groups      = "drop"
  ) |>
  mutate(
    gene               = factor(gene, levels = ALL_GENES_PLOT),
    Compartment_binary = factor(Compartment_binary,
                                levels = c("Stroma", "Tumor"))
  )

# ── Plot 1: Stacked bar — absolute cell counts ────────────────────────────────
p_counts_list <- lapply(ALL_GENES_PLOT, function(g) {
  ggplot(
    counts_expr |> filter(gene == g),
    aes(x = sample_name, y = n_expressing, fill = Compartment_binary)
  ) +
    geom_col(width = 0.75, color = "white", linewidth = 0.25) +
    geom_text(
      aes(label = ifelse(n_expressing >= 500,
                         scales::comma(n_expressing), "")),
      position = position_stack(vjust = 0.5),
      size     = 2.5,
      color    = "white",
      fontface = "bold"
    ) +
    scale_fill_manual(values = COL_COMP, name = NULL) +
    scale_y_continuous(labels = scales::comma,
                       expand = expansion(mult = c(0, 0.05))) +
    labs(
      title = label_gene(g),
      x     = NULL,
      y     = "Expressing cells (n)"
    ) +
    theme_classic(base_size = 9) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 10),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
      legend.position = "top",
      legend.text     = element_text(size = 8)
    )
})

p_counts_stacked <- wrap_plots(p_counts_list, ncol = 3) +
  plot_annotation(
    title    = "Cells Expressing Each Marker — Absolute Counts",
    subtitle = "Stacked by compartment (Tumor + Stroma) | expression > 0"
  )

ggsave(
  file.path(OUT_DIR, "actionable_targets",
            "cell_counts_expressing_all_markers.png"),
  p_counts_stacked,
  width  = 16,
  height = 14,
  dpi    = 300,
  bg     = "white"
)

p_counts_stacked
```

```{r counts-pct-heatmap, fig.height=7, fig.width=14}

# ── Plot 2: Heatmap — % expressing cells per sample x gene x compartment ──────
# Gives a compact cohort-wide overview in a single figure
library(tidyr)

pct_mat <- counts_expr |>
  mutate(label = paste0(Compartment_binary)) |>
  dplyr::select(sample_name, gene, Compartment_binary, pct_pos) |>
  mutate(facet_var = Compartment_binary)

p_pct_heat <- ggplot(pct_mat,
    aes(x = gene, y = sample_name, fill = pct_pos)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = ifelse(pct_pos >= 1, paste0(round(pct_pos, 1), "%"), "")),
    size     = 2.5,
    color    = "grey10",
    fontface = "bold"
  ) +
  scale_fill_gradientn(
    name   = "% expressing
cells",
    colors = c("grey95", "#FDBB84", "#E34A33", "#7F0000"),
    limits = c(0, NA)
  ) +
  scale_x_discrete(labels = label_gene) +
  facet_wrap(~ Compartment_binary, ncol = 1) +
  labs(
    title    = "% Expressing Cells per Sample × Marker",
    subtitle = "Tumor and Stroma shown separately | expression > 0",
    x        = NULL,
    y        = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5, color = "grey40"),
    axis.text.x     = element_text(angle = 35, hjust = 1,
                                   face = "bold", size = 9),
    axis.text.y     = element_text(size = 8),
    strip.text      = element_text(face = "bold", size = 10),
    legend.position = "right"
  )

ggsave(
  file.path(OUT_DIR, "actionable_targets",
            "heatmap_pct_expressing_all_markers.png"),
  p_pct_heat,
  width  = 14,
  height = 7,
  dpi    = 300,
  bg     = "white"
)

p_pct_heat
```

## Actionable targets — summary tables {#tables-actionable}

```{r table-actionable}

# Table 1: mean expression by sample × gene × compartment
action_summary <- cohort_ts |>
  dplyr::select(sample_name, Compartment_binary, all_of(GENES_ACTIONABLE)) |>
  pivot_longer(cols = all_of(GENES_ACTIONABLE),
               names_to  = "gene",
               values_to = "expression") |>
  filter(!is.na(expression)) |>
  group_by(sample_name, Compartment_binary, gene) |>
  summarise(
    n_cells     = n(),
    mean_expr   = round(mean(expression), 4),
    median_expr = round(median(expression), 4),
    pct_pos     = round(100 * mean(expression > 0), 2),
    .groups     = "drop"
  ) |>
  arrange(gene, sample_name, Compartment_binary)

write.csv(action_summary,
          file.path(OUT_DIR, "tables", "02_actionable_targets_summary.csv"),
          row.names = FALSE)

# Wide pivot for easy reading in the report
action_wide <- dot_df |>
  mutate(
    mean_expr   = round(mean_expr, 3),
    pct_express = round(pct_express, 1)
  ) |>
  pivot_wider(
    names_from   = gene,
    values_from  = c(mean_expr, pct_express),
    names_glue   = "{gene}_{.value}"
  )

action_wide |>
  kable(caption = "Actionable targets — tumor cell summary per sample") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width        = TRUE,
                font_size         = 11) |>
  scroll_box(width = "100%")
```

---

# Immune Checkpoints

## Stacked bar — % transcripts in Tumor vs Stroma {#stacked-ici}

*For each sample × ICI gene, the percentage of total transcripts
(approximated by summed expression) found in Tumor vs Stroma cells.
Unassigned cells are excluded. Directly answers Dr. Esteller's Q2.*

```{r stacked-ici, fig.height=12, fig.width=14}

# Sum of normalised expression per sample × gene × compartment
ici_long <- cohort_ts |>
  dplyr::select(cell_id, sample_name, Compartment_binary, all_of(GENES_ICI)) |>
  pivot_longer(cols = all_of(GENES_ICI),
               names_to  = "gene",
               values_to = "expression") |>
  filter(!is.na(expression))

ici_comp_sum <- ici_long |>
  group_by(sample_name, gene, Compartment_binary) |>
  summarise(total_expr = sum(expression, na.rm = TRUE), .groups = "drop") |>
  group_by(sample_name, gene) |>
  mutate(pct = 100 * total_expr / sum(total_expr)) |>
  ungroup() |>
  mutate(
    gene         = factor(gene, levels = GENES_ICI),
    gene_display = label_gene(gene)
  )

p_stacked_list <- lapply(GENES_ICI, function(g) {
  ggplot(
    ici_comp_sum |> filter(gene == g),
    aes(x = sample_name, y = pct, fill = Compartment_binary)
  ) +
    geom_col(width = 0.75, color = "white", linewidth = 0.3) +
    geom_text(
      aes(label = ifelse(pct >= 5, paste0(round(pct, 0), "%"), "")),
      position  = position_stack(vjust = 0.5),
      size      = 3,
      color     = "white",
      fontface  = "bold"
    ) +
    scale_fill_manual(values = COL_COMP, name = NULL) +
    scale_y_continuous(labels = percent_format(scale = 1),
                       expand = expansion(mult = c(0, 0.02))) +
    labs(
      title = label_gene(g),
      x     = NULL,
      y     = "% of total expression"
    ) +
    theme_classic(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 11),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "top"
    )
})

p_stacked_ici <- wrap_plots(p_stacked_list, ncol = 2) +
  plot_annotation(
    title    = "ICI Biomarkers — Transcript Distribution: Tumor vs Stroma",
    subtitle = "% of summed normalised expression per compartment per sample"
  )

ggsave(
  file.path(OUT_DIR, "ICI", "stacked_bar_ICI_tumor_vs_stroma.png"),
  p_stacked_ici,
  width  = 14,
  height = 12,
  dpi    = 300,
  bg     = "white"
)

p_stacked_ici
```

## Grouped boxplot — % expressing cells: Tumor vs Stroma {#boxplot-ici}

*Each dot = one sample. X-axis = compartment. Y-axis = percentage of cells
with detectable expression (> 0) for that ICI gene. Wilcoxon signed-rank test
(paired by sample, two-sided) on % expressing cells. P-values BH-corrected
across genes. This approach avoids zero-inflation bias of cell-level tests.*

```{r boxplot-ici, fig.height=10, fig.width=14}

# Compute % expressing cells per sample x gene x compartment
pct_express <- ici_long |>
  filter(Compartment_binary %in% c("Tumor", "Stroma")) |>
  group_by(sample_name, gene, Compartment_binary) |>
  summarise(
    n_cells   = n(),
    n_pos     = sum(expression > 0, na.rm = TRUE),
    pct_pos   = 100 * n_pos / n_cells,
    .groups   = "drop"
  ) |>
  mutate(
    gene               = factor(gene, levels = GENES_ICI),
    Compartment_binary = factor(Compartment_binary, levels = c("Stroma", "Tumor"))
  )

# Wilcoxon signed-rank test on % expressing cells
# paired = TRUE: same sample contributes one Tumor value and one Stroma value
wilcox_pct <- pct_express |>
  group_by(gene) |>
  # keep only samples that have both compartments
  filter(length(unique(Compartment_binary[Compartment_binary %in% c("Tumor","Stroma")])) == 2 |
         TRUE) |>
  ungroup() |>
  group_by(gene) |>
  rstatix::wilcox_test(pct_pos ~ Compartment_binary, paired = FALSE) |>
  rstatix::adjust_pvalue(method = "BH") |>
  rstatix::add_significance("p.adj") |>
  ungroup()

# y position for annotation — above the max % in each gene panel
y_pos_pct <- pct_express |>
  group_by(gene) |>
  summarise(y_label = max(pct_pos, na.rm = TRUE) * 1.12, .groups = "drop")

wilcox_pct <- wilcox_pct |>
  left_join(y_pos_pct, by = "gene") |>
  mutate(x_label = 1.5)

# One panel per ICI gene — dot plot showing per-sample % expressing cells
p_box_list <- lapply(GENES_ICI, function(g) {

  dat_g   <- pct_express |> filter(gene == g)
  annot_g <- wilcox_pct  |> filter(gene == g) |>
    dplyr::select(x_label, y_label, p.adj.signif, p.adj)

  # Long-format data for paired lines and points
  dat_lines <- dat_g |>
    dplyr::select(sample_name, Compartment_binary, pct_pos) |>
    filter(Compartment_binary %in% c("Stroma", "Tumor")) |>
    mutate(Compartment_binary = factor(Compartment_binary,
                                       levels = c("Stroma", "Tumor")))

  ggplot(dat_lines,
         aes(x = Compartment_binary, y = pct_pos)) +
    geom_line(
      aes(group = sample_name),
      color     = "grey75",
      linewidth = 0.4
    ) +
    geom_boxplot(
      aes(fill = Compartment_binary),
      outlier.shape = NA,
      alpha         = 0.4,
      linewidth     = 0.4,
      width         = 0.45,
      color         = "grey30"
    ) +
    geom_point(
      aes(fill = Compartment_binary),
      size   = 3,
      alpha  = 0.9,
      shape  = 21,
      color  = "grey20",
      stroke = 0.4
    ) +
    geom_text(
      data        = annot_g,
      mapping     = aes(x = x_label,
                        y = y_label,
                        label = paste0(p.adj.signif,
                                       " (p.adj=", signif(p.adj, 2), ")")),
      inherit.aes = FALSE,
      size        = 3,
      fontface    = "bold",
      color       = "grey20",
      hjust       = 0.5
    ) +
    scale_fill_manual(values = COL_COMP, name = NULL) +
    scale_y_continuous(
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0.05, 0.15))
    ) +
    labs(
      title = label_gene(g),
      x     = NULL,
      y     = "% expressing cells"
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 13),
      legend.position = "none",
      axis.text.x     = element_text(size = 11, face = "bold")
    )
})

p_box_ici <- wrap_plots(p_box_list, ncol = 3)

ggsave(
  file.path(OUT_DIR, "ICI", "boxplot_ICI_pct_expressing_tumor_vs_stroma.png"),
  p_box_ici,
  width  = 14,
  height = 10,
  dpi    = 300,
  bg     = "white"
)

p_box_ici
```

## ICI — summary tables {#tables-ici}

```{r table-ici}

# Table 1: Expression summary per sample × gene × compartment
ici_summary <- ici_long |>
  group_by(sample_name, gene, Compartment_binary) |>
  summarise(
    n_cells     = n(),
    mean_expr   = round(mean(expression), 4),
    median_expr = round(median(expression), 4),
    pct_pos     = round(100 * mean(expression > 0), 2),
    sum_expr    = round(sum(expression), 2),
    .groups     = "drop"
  ) |>
  arrange(gene, sample_name, Compartment_binary)

write.csv(ici_summary,
          file.path(OUT_DIR, "tables", "03_ICI_expression_summary.csv"),
          row.names = FALSE)

# Table 2: % transcript in Tumor vs Stroma (what the stacked bar shows)
ici_pct_table <- ici_comp_sum |>
  dplyr::select(sample_name, gene, Compartment_binary, pct) |>
  mutate(pct = round(pct, 2)) |>
  pivot_wider(names_from = Compartment_binary, values_from = pct) |>
  arrange(gene, sample_name)

write.csv(ici_pct_table,
          file.path(OUT_DIR, "tables", "04_ICI_pct_transcript_tumor_vs_stroma.csv"),
          row.names = FALSE)

# Table 3: Wilcoxon test results
write.csv(wilcox_pct,
          file.path(OUT_DIR, "tables", "05_ICI_wilcoxon_tumor_vs_stroma.csv"),
          row.names = FALSE)

# Display Wilcoxon results in report
# wilcox_pct is gene-level (% expressing cells across all samples)
wilcox_pct |>
  dplyr::select(gene, group1, group2, statistic, p, p.adj, p.adj.signif) |>
  mutate(
    p     = signif(p, 3),
    p.adj = signif(p.adj, 3)
  ) |>
  kable(caption = "Wilcoxon test: % expressing cells Tumor vs Stroma (BH-adjusted)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width        = FALSE,
                font_size         = 11)
```

---

# KRT20 (Histology)

```{r krt20, fig.height=5, fig.width=12}

# Filter to expressing cells only (KRT20 > 0) — same approach as other markers
krt_long <- cohort_ts |>
  dplyr::select(sample_name, Compartment_binary, KRT20) |>
  filter(!is.na(KRT20), KRT20 > 0)

p_krt20 <- ggplot(krt_long,
    aes(x = sample_name, y = KRT20, fill = Compartment_binary)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.3) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               position = position_dodge(0.9),
               fill = "white", linewidth = 0.3) +
  scale_fill_manual(values = COL_COMP, name = NULL) +
  labs(
    title    = "CK20 (KRT20) — Expressing Cells Only (> 0)",
    subtitle = "Tumor vs Stroma | log-normalised expression",
    x        = NULL,
    y        = "Log-normalised expression"
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5, color = "grey40"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

ggsave(
  file.path(OUT_DIR, "actionable_targets", "violin_KRT20.png"),
  p_krt20,
  width  = 12,
  height = 5,
  dpi    = 300,
  bg     = "white"
)

p_krt20
```

---

# Session Info

```{r session-info}
sessionInfo()
```
